---
title: "DATA3888 Interdisciplinary"
format: revealjs
editor: visual
---

## Loading 40732 Dataset and finding differentially expressed genes

```{r}
library(GEOquery)
```

```{r}
gse <- getGEO(filename = "data/GSE40732_series_matrix.txt.gz")
gse@annotation
```

```{r}
expr_matrix <- exprs(gse)
dim(expr_matrix)
head(expr_matrix[, 1:5])

pheno_data_40732 <- pData(gse)
head(pheno_data_40732, 80)

feature_data_40732 <- fData(gse)
head(feature_data_40732)
```

```{r}
# Step 1: Filter rows where rownames begin with "NM_"
nm_rows <- grep("^NM_", rownames(expr_matrix), value = TRUE)
expr_matrix_nm <- expr_matrix[nm_rows, ]

# Step 2: Trim last 6 characters from each rowname (accession suffix)
refseq_ids <- rownames(expr_matrix_nm)
```

```{r}
# Step 3: Load org.Hs.eg.db to map NM_ accessions to gene symbols
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  install.packages("BiocManager")
  BiocManager::install("org.Hs.eg.db")
}
```

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
```

```{r}
gene_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(refseq_ids),
  columns = c("REFSEQ", "SYMBOL"),
  keytype = "REFSEQ"
)

# Step 5: Convert expression matrix to data frame and join with gene symbols
expr_df <- as.data.frame(expr_matrix_nm)
expr_df$RefSeq <- rownames(expr_df)

# Merge the expression matrix with gene symbol annotations
expr_annotated <- merge(expr_df, gene_map, by.x = "RefSeq", by.y = "REFSEQ", all.x = TRUE)

# Optional: Move SYMBOL column to the front
expr_annotated <- expr_annotated[, c("SYMBOL", setdiff(names(expr_annotated), "SYMBOL"))]

# View result
head(expr_annotated)

```

```{r}
library(dplyr)

# Step 1: Remove rows with NA gene symbols
expr_annotated_clean <- expr_annotated[!is.na(expr_annotated$SYMBOL), ]

# Step 2: Group by SYMBOL and calculate row-wise mean for each gene
# We'll exclude the "RefSeq" column and average across all sample columns
expr_aggregated <- expr_annotated_clean %>%
  group_by(SYMBOL) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

expr_matrix_40732 <- as.data.frame(expr_aggregated)
rownames(expr_matrix_40732) <- expr_matrix_40732$SYMBOL
expr_matrix_40732$SYMBOL <- NULL

head(expr_matrix_40732)
```

```{r}
library(limma)

# Create group labels (Asthma vs Control)
group_40732 <- factor(ifelse(grepl("TRUE", pheno_data_40732$characteristics_ch1), "Asthma", "Control"))

# Create design matrix
design_40732 <- model.matrix(~0 + group_40732)
colnames(design_40732) <- levels(group_40732)

# Fit linear model
fit_40732 <- lmFit(expr_matrix_40732, design_40732)

# Create contrast: Asthma vs Control
contrast_40732 <- makeContrasts(Asthma - Control, levels = design_40732)
fit2_40732 <- contrasts.fit(fit_40732, contrast_40732)
fit2_40732 <- eBayes(fit2_40732)

# Get all DE results
de_results_40732 <- topTable(fit2_40732, number = Inf, sort.by = "P")

# Quick check
head(de_results_40732)

top100_40732 <- rownames(de_results_40732)[order(de_results_40732$P.Value)][1:100]
```

## Loading 67472 dataset and running random forest to confirm it can be merged

```{r}
library(GEOquery)
library(affy)

# (2) Extract the CEL files if you haven't already
untar("GSE67472_RAW.tar", exdir = "GSE67472/")

# (3) Read CEL files
cel_files <- list.celfiles("GSE67472/", full.names = TRUE)

# (4) Load raw data
raw_data <- ReadAffy(filenames = cel_files)

# (5) Normalize using RMA
norm_data <- rma(raw_data)

# (6) Extract expression matrix
expr_matrix_new <- exprs(norm_data)

# (7) Check result
dim(expr_matrix_new)
head(expr_matrix_new[, 1:5])

```

```{r}
library(hgu133plus2.db)

# Map probe IDs to Gene Symbols
gene_symbols <- mapIds(hgu133plus2.db,
                       keys = rownames(expr_matrix_new),
                       column = "SYMBOL",
                       keytype = "PROBEID",
                       multiVals = "first")

# Check mapping
cat("Number of probes with mapped gene symbols:", sum(!is.na(gene_symbols)), "\n")
head(gene_symbols, 20)
```

```{r}
# Step 1: Keep only rows that map to a gene symbol
expr_matrix_filtered <- expr_matrix_new[!is.na(gene_symbols), ]

# Step 2: Replace rownames with gene symbols
rownames(expr_matrix_filtered) <- gene_symbols[rownames(expr_matrix_filtered)]

# Step 3: Convert to data frame for aggregation
expr_df <- as.data.frame(expr_matrix_filtered)

# Step 4: Collapse rows with the same gene symbol by averaging across columns (samples)
expr_collapsed <- expr_df %>%
  group_by(GeneSymbol = rownames(expr_df)) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

# Step 5: Convert back to matrix-style format with SYMBOLs as rownames
expr_matrix_67472 <- as.data.frame(expr_collapsed)
rownames(expr_matrix_67472) <- expr_matrix_67472$GeneSymbol
expr_matrix_67472$GeneSymbol <- NULL

# --- PCA-based split for raw GSE67472 (before any ComBat) ---
pca_67472 <- prcomp(t(expr_matrix_67472), center = TRUE, scale. = TRUE)
split_labels_67472 <- ifelse(pca_67472$x[, 1] > 0, "batch1", "batch2")

samples_67472_batch1 <- rownames(pca_67472$x)[split_labels_67472 == "batch1"]
samples_67472_batch2 <- rownames(pca_67472$x)[split_labels_67472 == "batch2"]

raw_67472_batch1 <- expr_matrix_67472[, samples_67472_batch1]
raw_67472_batch2 <- expr_matrix_67472[, samples_67472_batch2]


colnames(expr_matrix_67472) <- sub("_.*", "", colnames(expr_matrix_67472))

# Optional: preview result
head(expr_matrix_67472)
length(rownames(expr_matrix_67472))


```

```{r}
gse_67472 <- getGEO(filename = "data/GSE67472_series_matrix.txt.gz")

# Extract phenotype data
pheno_data_67472 <- pData(gse_67472)

# Quick check
head(pheno_data_67472)

pheno_data_67472 <- pheno_data_67472[colnames(expr_matrix_67472), ]

# Check
head(pheno_data_67472)
```

```{r}
symbols_67472 <- rownames(expr_matrix_67472)
symbols_40732 <- rownames(expr_matrix_40732)
```

```{r}
# out of the 16947 cells that could be connected to a gene symbol in GSE67472, 163 could be matched to a gene symbol in GSE40732
shared_genes_40732 <- intersect(symbols_67472, symbols_40732)
length(intersect(symbols_67472, symbols_40732))
```

Now, building random forest model:

```{r}
library(randomForest)
library(caret) 

# Combine two phenotype tables (make sure rows = samples in combined expression matrix)
pheno_data_new <- rbind(
  data.frame(Sample = colnames(expr_matrix_67472), 
             Label = ifelse(pheno_data_67472$`disease state:ch1` == "asthma", "Asthma", "Control"),
             Dataset = "GSE67472"),
  data.frame(Sample = colnames(expr_matrix_40732), 
             Label = ifelse(grepl("TRUE", pheno_data_40732$characteristics_ch1), "Asthma", "Control"),
             Dataset = "GSE40732")
)

```

```{r}
train_y <- factor(ifelse(grepl("asthma", pheno_data_67472$characteristics_ch1.3, ignore.case = TRUE), 
                         "Asthma", "Control"))

test_y <- factor(ifelse(grepl("TRUE", pheno_data_40732$characteristics_ch1), "Asthma", "Control"))
```

```{r}
train_X <- as.data.frame(t(expr_matrix_67472))
test_X  <- as.data.frame(t(expr_matrix_40732))

# 3. Match sample order to phenotype data (assuming sample names are rownames)
rownames(train_X) <- rownames(pheno_data_67472)
rownames(test_X)  <- rownames(pheno_data_40732)

train_X <- train_X[rownames(pheno_data_67472), ]
test_X  <- test_X[rownames(pheno_data_40732), ]
```

```{r}
# Step 1: Get common genes
common_genes <- intersect(colnames(train_X), colnames(test_X))

# Step 2: Subset and reorder both datasets
train_X_aligned <- train_X[, common_genes]
test_X_aligned  <- test_X[, common_genes]

length(common_genes)
```

```{r}
# 4. Train the random forest
set.seed(3888)
rf_model <- randomForest(x = train_X_aligned, y = train_y, ntree = 500, importance = TRUE)

# 5. Predict and evaluate
rf_preds <- predict(rf_model, newdata = test_X_aligned)

conf_mat <- confusionMatrix(rf_preds, test_y)
print(conf_mat)

cat("Test Accuracy:", round(conf_mat$overall["Accuracy"], 3), "\n")
```

```{r}
table(rf_preds)
table(test_y)
table(train_y)
```

```{r}
merged_expr <- cbind(expr_matrix_67472[common_genes, ], expr_matrix_40732[common_genes, ])
merged_expr_t <- as.data.frame(t(merged_expr))
```

```{r}
dataset_labels <- c(rep("GSE67472", ncol(expr_matrix_67472)),
                    rep("GSE40732", ncol(expr_matrix_40732)))
```

```{r}
# Transpose BEFORE plotting so that each box represents a sample
merged_expr_samplewise <- as.data.frame(t(merged_expr))  # samples as rows

boxplot(as.data.frame(t(merged_expr_samplewise)),
        main = "Expression Distributions by Sample",
        las = 2,
        col = ifelse(dataset_labels == "GSE67472", "#66c2a5", "#fc8d62"),
        outline = FALSE,
        ylab = "Expression")

legend("topright", legend = c("GSE67472", "GSE40732"),
       fill = c("#66c2a5", "#fc8d62"), cex = 0.8)
```

```{r}
# Perform PCA
pca_res <- prcomp(merged_expr_t, scale. = TRUE)

# Prepare PCA data frame
pca_df <- data.frame(PC1 = pca_res$x[, 1],
                     PC2 = pca_res$x[, 2],
                     Dataset = dataset_labels)

# Plot
library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_minimal() +
  labs(title = "PCA of Combined Expression Data",
       x = "PC1",
       y = "PC2") +
  scale_color_manual(values = c("GSE67472" = "#66c2a5", "GSE40732" = "#fc8d62"))
```

## Assessing 67472 merge appropriateness

```{r}
library(limma)

# Create group labels (Asthma vs Control)
group_67472 <- factor(ifelse(grepl("asthma", pheno_data_67472$`disease state:ch1`, ignore.case = TRUE), "Asthma", "Control"))

# 2. Create design matrix
design_67472 <- model.matrix(~0 + group_67472)
colnames(design_67472) <- levels(group_67472)

# 3. Fit linear model on expr_matrix_67472
fit_67472 <- lmFit(expr_matrix_67472, design_67472)

# 4. Create contrast: Asthma vs Control
contrast_67472 <- makeContrasts(Asthma - Control, levels = design_67472)
fit2_67472 <- contrasts.fit(fit_67472, contrast_67472)
fit2_67472 <- eBayes(fit2_67472)

# 5. Get all DE results
de_results_67472 <- topTable(fit2_67472, number = Inf, sort.by = "P")

# Quick check
head(de_results_67472)
```

```{r}
sum(de_results_67472$adj.P.Val < 0.05)
```

```{r}
signif_de_67472 <- rownames(de_results_67472)[order(de_results_67472$P.Value)][1:1536]
top100_67472 <- rownames(de_results_67472)[order(de_results_67472$P.Value)][1:100]
top100_shared_both <- intersect(top100_40732, top100_67472)
print(top100_shared_both)
```

## Loading 143303 dataset and running random forest to confirm it can be merged

```{r}
cache_dir <- tempdir()  # Get the temporary directory
cache_files <- list.files(cache_dir, pattern = "GSE143303", full.names = TRUE)

# Remove any cached files related to GSE143304
file.remove(cache_files)
library(GEOquery)
gse143303 <- getGEO("GSE143303", GSEMatrix = TRUE)

```

```{r}
expr_matrix_gse143303 <- exprs(gse143303[[1]])  # Expression data from GSE143304
fdata_gse143303 <- fData(gse143303[[1]])  # Feature data from GSE143304
expr_df_gse143303 <- as.data.frame(expr_matrix_gse143303)

```

```{r}
expr_df_gse143303 <- as.data.frame(expr_matrix_gse143303)
expr_df_gse143303$Symbol <- fdata_gse143303$Symbol[match(rownames(expr_df_gse143303), rownames(fdata_gse143303))]
expr_df_gse143303 <- expr_df_gse143303[, c("Symbol", setdiff(names(expr_df_gse143303), "Symbol"))]
expr_df_gse143303$Symbol <- as.character(fdata_gse143303$Symbol)
expr_df_gse143303$Symbol <- as.character(expr_df_gse143303$Symbol)
```

```{r}
expr_df_gse143303_clean <- expr_df_gse143303[!is.na(expr_df_gse143303$Symbol), ]
 
```

```{r}
#head(expr_df_gse143303)


expr_aggregated_gse143303 <- expr_df_gse143303_clean %>%
  group_by(Symbol) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

expr_matrix_gse143303_cleaned <- as.data.frame(expr_aggregated_gse143303)

# Set 'Symbol' as row names and remove it from the data frame since it is now row names
rownames(expr_matrix_gse143303_cleaned) <- expr_matrix_gse143303_cleaned$Symbol
expr_matrix_gse143303_cleaned$Symbol <- NULL  # Remove 'Symbol' column

# View the cleaned expression matrix
head(expr_matrix_gse143303_cleaned)

```

```{r}
common_genes <- intersect(rownames(expr_matrix_gse143303_cleaned), rownames(expr_matrix_40732))

# Print the number of common genes
cat("Number of common genes:", length(common_genes), "\n")
cat("First few common genes:", head(common_genes), "\n")


```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)

expr_gse143303_subset <- expr_matrix_gse143303_cleaned[common_genes, ]
expr_gse47302_subset <- expr_matrix_40732[common_genes, ]

# Step 3: Combine the two datasets into a single long format data frame
# Add a Dataset column to differentiate the two datasets
expr_gse143303_long <- as.data.frame(expr_gse143303_subset) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(cols = -Gene, names_to = "Sample", values_to = "Expression") %>%
  mutate(Dataset = "GSE143303")

expr_gse47302_long <- as.data.frame(expr_gse47302_subset) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(cols = -Gene, names_to = "Sample", values_to = "Expression") %>%
  mutate(Dataset = "GSE47302")


```

```{r}
library(caret)
library(randomForest)
pData_gse143303 <- pData(gse143303[[1]])
train_data <- expr_matrix_gse143303_cleaned[common_genes, ]  # Training data (GSE143303)
test_data <- expr_matrix_40732[common_genes, ]    

common_genes <- intersect(rownames(train_data), rownames(test_data))

# Step 2: Subset the training and test data to include only the common genes
train_data_subset <- train_data[common_genes, ]  # Subset training data with common genes
test_data_subset <- test_data[common_genes, ]  


# Step 2: Prepare the training data with a response variable
# Assuming you have a target variable (e.g., "asthma") in GSE143303 for training
# For demonstration, let's assume we want to predict a binary outcome for asthma (0 = non-asthma, 1 = asthma)
train_labels <- ifelse(pData_gse143303$characteristics_ch1.1 == "disease status: Severe asthma", 1, 0)

cat("Number of train_labels:", length(train_labels), "\n")
cat("Number of columns in train_data:", ncol(train_data_subset), "\n")

train_data_long <- as.data.frame(t(train_data_subset))
colnames(train_data_long) <- rownames(train_data_subset)  # Assign gene names as column names

train_data_long$Label <- train_labels
# Confirm if 'Label' is in the columns
cat("Is 'Label' in train_data_long?", "Label" %in% colnames(train_data_long), "\n")


test_data_long <- as.data.frame(t(test_data_subset))
train_labels <- as.factor(train_labels)

# Check if train_labels is now a factor
cat("Is train_labels a factor?", is.factor(train_labels), "\n")



colnames(train_data_long) <- tolower(trimws(colnames(train_data_long)))
colnames(test_data_long) <- tolower(trimws(colnames(test_data_long)))
colnames(train_data_long) <- gsub("-", "_", colnames(train_data_long))  # Replace hyphens with underscores in training data
colnames(test_data_long) <- gsub("-", "_", colnames(test_data_long))



# Remove the 'ankhd1_eif4ebp3' gene from the test dataset (if it's present in test_data_long)
train_data_long <- train_data_long[, !colnames(train_data_long) %in% "ankhd1_eif4ebp3"]
test_data_long <- test_data_long[, !colnames(test_data_long) %in% "ankhd1_eif4ebp3"]

missing_in_test <- setdiff(colnames(train_data_long), colnames(test_data_long))
cat("Genes in train_data_long but not in test_data_long:", missing_in_test, "\n")

# Find genes that are in test_data_long but not in train_data_long
missing_in_train <- setdiff(colnames(test_data_long), colnames(train_data_long))
cat("Genes in test_data_long but not in train_data_long:", missing_in_train, "\n")



rf_model <- randomForest(label ~ ., data = train_data_long, ntree = 500)  # ntree = 500 trees

# Step 4: Test the model on GSE47302 (Test data)
#test_data_long <- as.data.frame(t(test_data_subset))
predictions <- predict(rf_model, newdata = test_data_long)


# Convert predictions to binary (0 or 1)
predicted_labels <- ifelse(predictions > 0.5, 1, 0)

# Step 5: Evaluate model performance (e.g., accuracy, confusion matrix)
# Assuming you have true labels for the test set
test_labels <- ifelse(pheno_data_40732$characteristics_ch1 == "asthma: TRUE", 1, 0)

# Create confusion matrix
conf_matrix <- confusionMatrix(as.factor(predicted_labels), as.factor(test_labels))

# Print evaluation metrics
print(conf_matrix)
```

```{r}
train_data_long_pca <- t(train_data_long)  # Transpose the training data
test_data_long_pca <- t(test_data_long)
train_data_long_pca <- train_data_long_pca[rownames(train_data_long_pca) != "label", ]

cat("Dimensions of transposed train data:", dim(train_data_long_pca), "\n")
cat("Dimensions of transposed test data:", dim(test_data_long_pca), "\n")

combined_data <- cbind(train_data_long_pca, test_data_long_pca)

# Check the dimensions of combined_data to verify
cat("Dimensions of combined_data:", dim(combined_data), "\n")
```

```{r}
combined_data_pca<- combined_data


pca <- prcomp(t(combined_data_pca), center = TRUE, scale. = TRUE)

# Create a PCA plot (example)
pca_df <- data.frame(pca$x)  # Extract PCA results

# Add dataset labels (for visualization purposes)
pca_df$Dataset <- c(rep("GSE143303", ncol(train_data_long_pca)), rep("GSE47302", ncol(test_data_long_pca)))

cat("Dimensions of pca_df:", dim(pca_df), "\n")
cat("Length of Dataset column:", length(pca_df$Dataset), "\n")

# Plot PCA
library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 3) +
  labs(title = "PCA of Gene Expression",
       x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```

## Assessing 143303 merge appropriateness

```{r}
library(limma)

# Expression matrix = expr_matrix_gse143303_cleaned
# Phenotype data:
pheno_data_gse143303 <- pData(gse143303[[1]])

# Clean sample names if needed (GSM IDs match)
colnames(expr_matrix_gse143303_cleaned) <- sub("_.*", "", colnames(expr_matrix_gse143303_cleaned))
pheno_data_gse143303 <- pheno_data_gse143303[colnames(expr_matrix_gse143303_cleaned), ]

# Create group labels (Severe asthma vs others)
group_143303_cleaned <- factor(ifelse(pheno_data_gse143303$characteristics_ch1.1 == "disease status: Severe asthma", "Asthma", "Control"))

# Build design matrix
design_143303_cleaned <- model.matrix(~0 + group_143303_cleaned)
colnames(design_143303_cleaned) <- levels(group_143303_cleaned)

# Fit linear model
fit_143303_cleaned <- lmFit(expr_matrix_gse143303_cleaned, design_143303_cleaned)

# Create contrast: Asthma vs Control
contrast_143303_cleaned <- makeContrasts(Asthma - Control, levels = design_143303_cleaned)
fit2_143303_cleaned <- contrasts.fit(fit_143303_cleaned, contrast_143303_cleaned)
fit2_143303_cleaned <- eBayes(fit2_143303_cleaned)

# Extract DE results
de_results_143303 <- topTable(fit2_143303_cleaned, number = Inf, sort.by = "P")

# Quick view
head(de_results_143303)

signif_de_143303 <- rownames(de_results_143303)[order(de_results_143303$P.Value)][1:2043]

# Top 100 genes (gene symbols)
top100_143303 <- rownames(de_results_143303)[order(de_results_143303$P.Value)][1:100]

# View
print(top100_143303)

```

```{r}
print(top100_40732)
```

```{r}
shared_top100_genes_143303 <- intersect(top100_40732, top100_143303)

# View shared genes
print(shared_top100_genes_143303)

# How many shared?
length(shared_top100_genes_143303)
```

```{r}
shared_signif_143303x67472 <- intersect(signif_de_143303, signif_de_67472)

# View shared genes
print(shared_signif_143303x67472)

# How many shared?
length(shared_signif_143303x67472)
```

### Kimi's stuff

```{r}
gene_means <- rowMeans(combined_data, na.rm = TRUE)

# Get the top 100 genes with the highest mean expression
top_100_genes <- names(sort(gene_means, decreasing = TRUE))[1:100]

# Subset the combined data to include only the top 100 genes
top_100_data <- combined_data[top_100_genes, ]

# Reshape the data to long format for ggplot
library(tidyr)
top_100_data_long <- as.data.frame(top_100_data) %>%
  rownames_to_column("Gene") %>%
  pivot_longer(cols = -Gene, names_to = "Sample", values_to = "Expression") %>%
  mutate(Dataset = ifelse(Sample %in% colnames(train_data_long), "GSE143303", "GSE47302"))

```

## Loading 40888

```{r}
gse_40888 <- getGEO(filename = "data/GSE40888_series_matrix.txt.gz")
```

```{r}
expr_matrix_40888 <- exprs(gse_40888)
dim(expr_matrix_40888)
head(expr_matrix_40888[, 1:5])

pheno_data_40888 <- pData(gse_40888)
head(pheno_data_40888, 20)

feature_data_40888 <- fData(gse_40888)
head(feature_data_40888, 20)
```

```{r}
colnames(feature_data_40888)
```

```{r}
# Extract probe ID → SYMBOL correspondence (from gene_assignment)
library(dplyr)
library(stringr)

probe2symbol <- feature_data_40888 %>%
  mutate(SYMBOL = str_split_fixed(gene_assignment, " // ", 3)[,2]) %>%
  select(ID, SYMBOL) %>%
  filter(!is.na(SYMBOL), SYMBOL != "")
```

```{r}
# Combining expression matrices with SYMBOL
expr_df_40888 <- as.data.frame(expr_matrix_40888)
expr_df_40888$ID <- rownames(expr_df_40888)

expr_annotated_40888 <- merge(expr_df_40888, probe2symbol, by = "ID")

# Aggregate the expression values of multiple probes for the same gene 
expr_clean_40888 <- expr_annotated_40888 %>%
  group_by(SYMBOL) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))

expr_matrix_40888_clean <- as.data.frame(expr_clean_40888)
rownames(expr_matrix_40888_clean) <- expr_matrix_40888_clean$SYMBOL
expr_matrix_40888_clean$SYMBOL <- NULL

# --- PCA-based split for raw GSE40888 (before any ComBat) ---
pca_40888 <- prcomp(t(expr_matrix_40888_clean), center = TRUE, scale. = TRUE)
split_labels_40888 <- ifelse(pca_40888$x[, 1] > 0, "batch1", "batch2")

samples_40888_batch1 <- rownames(pca_40888$x)[split_labels_40888 == "batch1"]
samples_40888_batch2 <- rownames(pca_40888$x)[split_labels_40888 == "batch2"]

raw_40888_batch1 <- expr_matrix_40888_clean[, samples_40888_batch1]
raw_40888_batch2 <- expr_matrix_40888_clean[, samples_40888_batch2]

# === Merge raw (pre-ComBat) expression matrix for 7 datasets ===

# Step 1: Get common genes across all 7 datasets
common_genes_all <- Reduce(intersect, list(
  rownames(expr_matrix_64913_clean),
  rownames(raw_40888_batch1),
  rownames(raw_40888_batch2),
  rownames(expr_matrix_gse143303_cleaned),
  rownames(raw_67472_batch1),
  rownames(raw_67472_batch2),
  rownames(expr_matrix_40732)
))

# Step 2: Combine into one matrix (raw, uncorrected)
merged_raw_7 <- cbind(
  expr_matrix_64913_clean[common_genes_all, ],
  raw_40888_batch1[common_genes_all, ],
  raw_40888_batch2[common_genes_all, ],
  expr_matrix_gse143303_cleaned[common_genes_all, ],
  raw_67472_batch1[common_genes_all, ],
  raw_67472_batch2[common_genes_all, ],
  expr_matrix_40732[common_genes_all, ]
)

# Step 3: Export to CSV
write.csv(merged_raw_7, "expression_7datasets_preCombat.csv")


expr_matrix_40888_clean
```

```{r}
symbols_40888 <- rownames(expr_matrix_40888_clean)
```

## Assessing 40888 for merge appropriateness

```{r}
library(limma)

# Create group labels (Severe asthma vs others)
group_40888_cleaned <- factor(ifelse(grepl("asthmatic", pheno_data_40888$`characteristics_ch1.3`, ignore.case = TRUE), "Asthma", "Control"))

# Build design matrix
design_40888_cleaned <- model.matrix(~0 + group_40888_cleaned)
colnames(design_40888_cleaned) <- levels(group_40888_cleaned)

# Fit linear model
fit_40888_cleaned <- lmFit(expr_matrix_40888_clean, design_40888_cleaned)

# Create contrast: Asthma vs Control
contrast_40888_cleaned <- makeContrasts(Asthma - Control, levels = design_40888_cleaned)
fit2_40888_cleaned <- contrasts.fit(fit_40888_cleaned, contrast_40888_cleaned)
fit2_40888_cleaned <- eBayes(fit2_40888_cleaned)

# Extract DE results
de_results_40888 <- topTable(fit2_40888_cleaned, number = Inf, sort.by = "P")

# Quick view
head(de_results_40888)

signif_de_40888 <- rownames(de_results_40888)[order(de_results_40888$P.Value)][1:1029]

# Top 100 genes (gene symbols)
top100_40888 <- rownames(de_results_40888)[order(de_results_40888$P.Value)][1:100]

# View
print(top100_40888)

```

```{r}
shared_top100_genes_40888 <- intersect(top100_40732, top100_40888)

# View shared genes
print(shared_top100_genes_40888)

# How many shared?
length(shared_top100_genes_40888)
```

```{r}
shared_signif_143303x67472x40888 <- intersect(intersect(signif_de_143303, signif_de_67472), signif_de_40888)

# View shared genes
print(shared_signif_143303x67472x40888)

# How many shared?
length(shared_signif_143303x67472x40888)
```

```{r}
shared_signif_67472x40888 <- intersect(signif_de_40888, signif_de_67472)

# View shared genes
print(shared_signif_67472x40888)

# How many shared?
length(shared_signif_67472x40888)
```

```{r}
shared_signif_143303x40888 <- intersect(signif_de_40888, signif_de_143303)

# View shared genes
print(shared_signif_143303x40888)

# How many shared?
length(shared_signif_143303x40888)
```

## Loading 188424

```{r}
# --- Step 2: Load GSE188424 dataset ---
gse188424 <- getGEO(filename = "data/GSE188424_series_matrix.txt.gz")
```

```{r}
# --- Step 3: Extract expression matrix, phenotype and feature data ---
expr_matrix_188424 <- exprs(gse188424)
pheno_data_188424 <- pData(gse188424)
feature_data_188424 <- fData(gse188424)

head(expr_matrix_188424,20)
head(pheno_data_188424,20)
head(feature_data_188424,20)
```

```{r}
# Load GEO platform annotation
platform_id <- annotation(gse188424)  # Get platform ID, e.g., GPLxxxx
platform <- getGEO(platform_id)
feature_data <- Table(platform)

# Extract probe ID and gene symbol mapping
feature_map_188424 <- feature_data[, c("ID", "Symbol")]
colnames(feature_map_188424) <- c("probe_id", "symbol")

# Step 3: Map probe IDs in expr_matrix_188424 to gene symbols 
expr_df_188424 <- as.data.frame(expr_matrix_188424)
expr_df_188424$probe_id <- rownames(expr_df_188424)

expr_annotated_188424 <- merge(expr_df_188424, feature_map_188424, by = "probe_id", all.x = TRUE)

# Move 'symbol' column to the front
expr_annotated_188424 <- expr_annotated_188424[, c("symbol", setdiff(names(expr_annotated_188424), "symbol"))]

# Remove rows with missing gene symbols 
expr_annotated_188424_clean <- expr_annotated_188424[!is.na(expr_annotated_188424$symbol), ]

# Aggregate expression values by gene symbol 
library(dplyr)
expr_aggregated_188424 <- expr_annotated_188424_clean %>%
  group_by(symbol) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))

# Format back to matrix 
expr_matrix_188424_cleaned <- as.data.frame(expr_aggregated_188424)
rownames(expr_matrix_188424_cleaned) <- expr_matrix_188424_cleaned$symbol
expr_matrix_188424_cleaned$symbol <- NULL



```

```{r}
# --- Step 5: Remove NA symbols and aggregate by symbol ---
expr_annotated_188424_clean <- expr_annotated_188424[!is.na(expr_annotated_188424$symbol), ]

expr_aggregated_188424 <- expr_annotated_188424_clean %>%
  group_by(symbol) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

```{r}
# --- Step 6: Final training expression matrix ---
expr_matrix_188424_clean <- as.data.frame(expr_aggregated_188424)
rownames(expr_matrix_188424_clean) <- expr_matrix_188424_clean$symbol
expr_matrix_188424_clean$symbol <- NULL

head(expr_matrix_188424_clean, 20)

```

```{r}
# --- Step 7: Prepare labels (Controlled / Uncontrolled) ---
title_labels <- pheno_data_188424$title
condition <- gsub(" .*", "", title_labels)  
table(condition)

# Align condition with training samples
condition <- condition[1:ncol(expr_matrix_188424_clean)]
selected <- condition %in% c("Controlled", "Uncontrolled")
condition <- condition[1:ncol(expr_matrix_188424_clean)]
train_y <- ifelse(condition[selected] == "Controlled", 1, 0)
train_y <- as.factor(train_y)
```

```{r}
# --- Step 8: Model training ---
# --- Step 8: Prepare train_X and train_y and run model ---
# Transpose expression matrix so rows = samples, columns = genes
train_X <- as.data.frame(t(expr_matrix_188424_clean[, selected]))

# Make sure labels are a factor
train_y <- as.factor(ifelse(condition[selected] == "Controlled", 1, 0))

# Confirm dimensions match
stopifnot(nrow(train_X) == length(train_y))

# Train the random forest
set.seed(3888)
rf_model <- randomForest(x = train_X, y = train_y, ntree = 500, importance = TRUE)
```

```{r}
# Align common genes
shared_genes_188424 <- intersect(rownames(expr_matrix_188424_clean), rownames(expr_matrix_40732))

length(shared_genes_188424)
```

```{r}
# Merge the two datasets
combined_data <- cbind(
  expr_matrix_188424_cleaned[common_genes, ],
  expr_matrix_40732[common_genes, ]
)

# Transpose the matrix 
# Rows become samples, columns become genes
combined_data_t <- t(combined_data)

# Standardize the data 
# For each gene (column), scale to mean = 0 and variance = 1
combined_data_scaled <- scale(combined_data_t)
```

```{r}
sum(is.na(combined_data_scaled))       # Should be 0
sum(is.infinite(combined_data_scaled)) # Should be 0
```

```{r}
# Remove genes (columns) with any NA values
combined_data_scaled_clean <- combined_data_scaled[, colSums(is.na(combined_data_scaled)) == 0]

# Perform PCA 
pca_result <- prcomp(combined_data_scaled_clean, center = TRUE, scale. = TRUE)

# Organize PCA results into a dataframe 
pca_df <- data.frame(pca_result$x)

# Add dataset labels for each sample 
pca_df$Dataset <- ifelse(
  rownames(pca_df) %in% colnames(expr_matrix_188424_cleaned),
  "GSE188424",
  "GSE40732"
)

# Plot PCA 
library(ggplot2)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 2) +
  labs(title = "PCA of Combined Expression Data",
       x = "PC1", y = "PC2") +
  theme_minimal()


```

## Assessing 188424 for merge appropriateness

```{r}
library(limma)


# Create group labels (Severe asthma vs others)
group_188424_cleaned <- factor(ifelse(pheno_data_188424$description == "Uncontrolled", "Asthma", "Control"))

# Build design matrix
design_188424_cleaned <- model.matrix(~0 + group_188424_cleaned)
colnames(design_188424_cleaned) <- levels(group_188424_cleaned)

# Fit linear model
fit_188424_cleaned <- lmFit(expr_matrix_188424_clean, design_188424_cleaned)

# Create contrast: Asthma vs Control
contrast_188424_cleaned <- makeContrasts(Asthma - Control, levels = design_188424_cleaned)
fit2_188424_cleaned <- contrasts.fit(fit_188424_cleaned, contrast_188424_cleaned)
fit2_188424_cleaned <- eBayes(fit2_188424_cleaned)

# Extract DE results
de_results_188424 <- topTable(fit2_188424_cleaned, number = Inf, sort.by = "P")

# Quick view
head(de_results_188424)

signif_de_188424 <- rownames(de_results_188424)[order(de_results_188424$P.Value)][1:38]

# Top 100 genes (gene symbols)
top100_188424 <- rownames(de_results_188424)[order(de_results_188424$P.Value)][1:100]

# View
print(top100_188424)

```

## Loading 64913

```{r}
# 1. Load GSE64913
gse_64913 <- getGEO(filename = "data/GSE64913_series_matrix.txt.gz")

# 2. Extract expression matrix
expr_matrix_64913 <- exprs(gse_64913)
dim(expr_matrix_64913)
head(expr_matrix_64913[, 1:5])

# 3. Extract phenotype (sample info)
pheno_data_64913 <- pData(gse_64913)
head(pheno_data_64913, 100)

# 4. Extract feature data (probe annotations)
feature_data_64913 <- fData(gse_64913)
head(feature_data_64913, 20)
```

```{r}
colnames(feature_data_64913)
```

```{r}
if ("Gene Symbol" %in% colnames(feature_data_64913)) {
  probe2symbol_64913 <- feature_data_64913 %>%
    select(ID, SYMBOL = `Gene Symbol`) %>%
    filter(!is.na(SYMBOL), SYMBOL != "")
} else if ("gene_assignment" %in% colnames(feature_data_64913)) {
  # Fallback for gene_assignment format (like 40888)
  probe2symbol_64913 <- feature_data_64913 %>%
    mutate(SYMBOL = str_split_fixed(gene_assignment, " // ", 3)[,2]) %>%
    select(ID, SYMBOL) %>%
    filter(!is.na(SYMBOL), SYMBOL != "")
}
```

```{r}
# Merge expression data with SYMBOL
expr_df_64913 <- as.data.frame(expr_matrix_64913)
expr_df_64913$ID <- rownames(expr_df_64913)

expr_annotated_64913 <- merge(expr_df_64913, probe2symbol_64913, by = "ID")

# Collapse multiple probes per gene by averaging
expr_clean_64913 <- expr_annotated_64913 %>%
  group_by(SYMBOL) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))

expr_matrix_64913_clean <- as.data.frame(expr_clean_64913)
rownames(expr_matrix_64913_clean) <- expr_matrix_64913_clean$SYMBOL
expr_matrix_64913_clean$SYMBOL <- NULL

# View cleaned matrix
head(expr_matrix_64913_clean)
```

```{r}
symbols_64913 <- rownames(expr_matrix_64913_clean)

```

```{r}
shared_genes_64913 <- intersect(symbols_64913, symbols_40732)

# View
length(shared_genes_64913)
```

## Assessing 64913

```{r}
library(limma)

# Create group labels for GSE64913
# Check which field separates asthma vs control in pheno_data_64913
# (Example below assumes 'disease status:ch1' or similar field)

group_64913_cleaned <- factor(ifelse(grepl("Asthmatic", pheno_data_64913$characteristics_ch1.1, ignore.case = TRUE), "Asthma", "Control"))

# Build design matrix
design_64913_cleaned <- model.matrix(~0 + group_64913_cleaned)
colnames(design_64913_cleaned) <- levels(group_64913_cleaned)

# Fit linear model
fit_64913_cleaned <- lmFit(expr_matrix_64913_clean, design_64913_cleaned)

# Create contrast: Asthma vs Control
contrast_64913_cleaned <- makeContrasts(Asthma - Control, levels = design_64913_cleaned)
fit2_64913_cleaned <- contrasts.fit(fit_64913_cleaned, contrast_64913_cleaned)
fit2_64913_cleaned <- eBayes(fit2_64913_cleaned)

# Extract DE results
de_results_64913 <- topTable(fit2_64913_cleaned, number = Inf, sort.by = "P")

# View results
head(de_results_64913)
sum(de_results_64913$adj.P.Val < 0.05)

signif_de_64913 <- rownames(de_results_64913)[order(de_results_64913$P.Value)][1:915]

# Top 100 genes (gene symbols)
top100_64913 <- rownames(de_results_64913)[order(de_results_64913$P.Value)][1:100]

```

```{r}
# 64913 x 67472
shared_signif_64913x67472 <- intersect(signif_de_64913, signif_de_67472)
cat("64913 x 67472 shared genes:", length(shared_signif_64913x67472), "\n")

# 64913 x 143303
shared_signif_64913x143303 <- intersect(signif_de_64913, signif_de_143303)
cat("64913 x 143303 shared genes:", length(shared_signif_64913x143303), "\n")

# 64913 x 40888
shared_signif_64913x40888 <- intersect(signif_de_64913, signif_de_40888)
cat("64913 x 40888 shared genes:", length(shared_signif_64913x40888), "\n")

# 64913 x 40888 x 143303
shared_signif_64913x40888x143303 <- intersect(signif_de_64913, intersect(signif_de_40888, signif_de_143303))
cat("64913 x 40888 x 143303 shared genes:", length(shared_signif_64913x40888x143303), "\n")
```

```{r}
# 64913 x 40888 x 143303
shared_64913x40888x143303 <- intersect(rownames(expr_matrix_64913_clean), intersect(rownames(expr_matrix_40888_clean),rownames(expr_matrix_gse143303_cleaned)))
cat("64913 x 40888 x 143303 shared genes:", length(shared_64913x40888x143303), "\n")

# 64913 x 40888 x 143303 x 67472
shared_64913x40888x143303x67472 <- intersect(intersect(rownames(expr_matrix_64913_clean), rownames(expr_matrix_67472)), intersect(rownames(expr_matrix_40888_clean),rownames(expr_matrix_gse143303_cleaned)))
cat("64913 x 40888 x 143303 x 67472 shared genes:", length(shared_64913x40888x143303x67472), "\n")

# 64913 x 40888 x 143303 x 67472 x 40732
shared_64913x40888x143303x67472x40732 <- intersect(intersect(rownames(expr_matrix_40732), intersect(rownames(expr_matrix_64913_clean), rownames(expr_matrix_67472))), intersect(rownames(expr_matrix_40888_clean),rownames(expr_matrix_gse143303_cleaned)))
cat("64913 x 40888 x 143303 x 67472 x 40732 shared genes:", length(shared_64913x40888x143303x67472x40732), "\n")
```

```{r}
common_genes_train <- Reduce(intersect, list(
  rownames(expr_matrix_64913_clean),
  rownames(expr_matrix_40888_clean),
  rownames(expr_matrix_gse143303_cleaned),
  rownames(expr_matrix_67472)
))
length(common_genes_train)
```

```{r}
merged_train <- cbind(
  expr_matrix_64913_clean[common_genes_train, ],
  expr_matrix_40888_clean[common_genes_train, ],
  expr_matrix_gse143303_cleaned[common_genes_train, ],
  expr_matrix_67472[common_genes_train, ]
)

# Transpose for sample-wise rows
merged_train_t <- t(merged_train)
```

```{r}
# --- Adjust batch effects using ComBat ---
library(sva)

# 1. Merge train and test datasets with common genes
common_genes_all <- Reduce(intersect, list(
  rownames(expr_matrix_64913_clean),
  rownames(expr_matrix_40888_clean),
  rownames(expr_matrix_gse143303_cleaned),
  rownames(expr_matrix_67472),
  rownames(expr_matrix_40732)  # include test set!
))

merged_full <- cbind(
  expr_matrix_64913_clean[common_genes_all, ],
  expr_matrix_40888_clean[common_genes_all, ],
  expr_matrix_gse143303_cleaned[common_genes_all, ],
  expr_matrix_67472[common_genes_all, ],
  expr_matrix_40732[common_genes_all, ]
)

# 2. Construct batch labels
full_batch_labels <- c(
  rep("GSE64913", ncol(expr_matrix_64913_clean)),
  rep("GSE40888", ncol(expr_matrix_40888_clean)),
  rep("GSE143303", ncol(expr_matrix_gse143303_cleaned)),
  rep("GSE67472", ncol(expr_matrix_67472)),
  rep("GSE40732", ncol(expr_matrix_40732))
)

# 3. Apply ComBat
combat_corrected_full <- ComBat(dat = merged_full, batch = full_batch_labels, par.prior = TRUE)

# 4. Split into train and test
n_train <- length(full_batch_labels) - ncol(expr_matrix_40732)
combat_train <- combat_corrected_full[, 1:n_train]
combat_test  <- combat_corrected_full[, (n_train + 1):ncol(combat_corrected_full)]

# 5. Transpose train matrix for randomForest
combat_train_t <- as.data.frame(t(combat_train))

```

```{r}

pheno_data_64913 <- pheno_data_64913[colnames(expr_matrix_64913_clean), ]
pheno_data_40888 <- pheno_data_40888[colnames(expr_matrix_40888_clean), ]
pheno_data_gse143303 <- pheno_data_gse143303[colnames(expr_matrix_gse143303_cleaned), ]
pheno_data_67472 <- pheno_data_67472[colnames(expr_matrix_67472), ]

labels_64913 <- ifelse(grepl("Asthmatic", pheno_data_64913$characteristics_ch1.1, ignore.case = TRUE), "Asthma", "Control")
labels_40888 <- ifelse(grepl("asthmatic", pheno_data_40888$`characteristics_ch1.3`, ignore.case = TRUE), "Asthma", "Control")
labels_143303 <- ifelse(pheno_data_gse143303$characteristics_ch1.1 == "disease status: Severe asthma", "Asthma", "Control")
labels_67472 <- ifelse(grepl("asthma", pheno_data_67472$`disease state:ch1`, ignore.case = TRUE), "Asthma", "Control")

train_labels <- factor(c(labels_64913, labels_40888, labels_143303, labels_67472))

```

```{r}
dim(combat_train_t)      # Should be: samples x genes
length(train_labels)    # Should equal number of rows in combat_expr_t

```

```{r}
# Use combat_corrected_full to find common genes with the test set
common_genes_test <- intersect(rownames(expr_matrix_40732), rownames(combat_corrected_full))

# Subset the corrected matrix for test samples (i.e., GSE40732 samples)
test_expr_40732 <- combat_corrected_full[common_genes_test, colnames(expr_matrix_40732)]

```

```{r}
test_expr_40732_t <- t(test_expr_40732)
```

```{r}
test_labels <- factor(ifelse(grepl("TRUE", pheno_data_40732$characteristics_ch1), "Asthma", "Control"))

```

```{r}
# Step 1: Get genes shared across all datasets (including GSE40732)
common_genes_all <- Reduce(intersect, list(
  rownames(expr_matrix_64913_clean),
  rownames(expr_matrix_40888_clean),
  rownames(expr_matrix_gse143303_cleaned),
  rownames(expr_matrix_67472),
  rownames(expr_matrix_40732)
))

# Step 2: Combine ALL uncorrected expression matrices
merged_full <- cbind(
  expr_matrix_64913_clean[common_genes_all, ],
  expr_matrix_40888_clean[common_genes_all, ],
  expr_matrix_gse143303_cleaned[common_genes_all, ],
  expr_matrix_67472[common_genes_all, ],
  expr_matrix_40732[common_genes_all, ]
)

# Step 3: Construct full batch labels (must match sample order in merged_full)
full_batch_labels <- c(
  rep("GSE64913", ncol(expr_matrix_64913_clean)),
  rep("GSE40888", ncol(expr_matrix_40888_clean)),
  rep("GSE143303", ncol(expr_matrix_gse143303_cleaned)),
  rep("GSE67472", ncol(expr_matrix_67472)),
  rep("GSE40732", ncol(expr_matrix_40732))
)

library(sva)

# --- Separate training and test data ---
n_train <- length(full_batch_labels) - ncol(expr_matrix_40732)
merged_train <- merged_full[, 1:n_train]
train_batch_labels <- full_batch_labels[1:n_train]

# Step 1: Run ComBat on training data
combat_train <- ComBat(
  dat = merged_train,
  batch = train_batch_labels,
  par.prior = TRUE
)

# Step 2: Estimate mean and variance from training data (manually)
train_means <- rowMeans(combat_train, na.rm = TRUE)
train_sds <- apply(combat_train, 1, sd, na.rm = TRUE)

# Step 3: Save the estimates as a frozen model
combat_frozen_model <- list(
  stand.mean = train_means,
  stand.sd = train_sds
)

saveRDS(combat_frozen_model, file = "combat_frozen_model_basic.rds")

# Step 4: Run ComBat once across the full matrix
combat_corrected_full <- ComBat(dat = merged_full, batch = full_batch_labels, par.prior = TRUE)


saveRDS(combat_corrected_full, file = "frozen_combat_corrected_expr_primary_model.rds")
```

```{r}
combat_test_t <- as.data.frame(t(combat_test))  # transpose so samples are rows

# Step 1: Align test set with training genes
train_genes <- colnames(combat_train_t)
test_genes  <- colnames(combat_test_t)

# Step 2: Find common genes
common_genes <- intersect(train_genes, test_genes)

# Step 3: Subset both training and test data to shared genes
combat_train_t_aligned <- combat_train_t[, common_genes]
combat_test_t_aligned  <- combat_test_t[, common_genes]

# (Optional sanity check)
stopifnot(all(colnames(combat_train_t_aligned) == colnames(combat_test_t_aligned)))

```

```{r}
# Split corrected data into train and test sets
n_train <- length(full_batch_labels) - ncol(expr_matrix_40732)

combat_train <- combat_corrected_full[, 1:n_train]
combat_test  <- combat_corrected_full[, (n_train + 1):ncol(combat_corrected_full)]
```

```{r}
table(train_labels)
```

```{r}
rf_model <- randomForest(x = combat_train_t_aligned, y = train_labels, ntree = 500)

# Predict on test data
rf_preds <- predict(rf_model, newdata = combat_test_t_aligned)

# Evaluate
conf_matrix <- confusionMatrix(rf_preds, test_labels)
print(conf_matrix)

cat("Test Accuracy:", round(conf_matrix$overall["Accuracy"], 3), "\n")

```

```{r}
expr_for_boxplot <- as.data.frame(t(combat_train))  # use updated combat_train
sample_ids <- colnames(combat_train)
```

```{r}
# Align gene rows across train and test
common_genes <- intersect(rownames(combat_train), rownames(combat_test))

combat_train_aligned <- combat_train[common_genes, ]
combat_test_aligned  <- combat_test[common_genes, ]

# Define dataset labels
batch_labels <- c(
  rep("GSE64913", length(labels_64913)),
  rep("GSE40888", length(labels_40888)),
  rep("GSE143303", length(labels_143303)),
  rep("GSE67472", length(labels_67472))
)
combined_dataset_labels <- c(
  batch_labels,
  rep("GSE40732", ncol(combat_test_aligned))
)

# Prepare expression matrix and metadata
all_expr <- cbind(combat_train_aligned, combat_test_aligned)
expr_for_boxplot <- as.data.frame(t(all_expr))
expr_for_boxplot$Sample <- rownames(expr_for_boxplot)
expr_for_boxplot$Dataset <- factor(combined_dataset_labels)

# Reshape
library(tidyr)
library(dplyr)
library(ggplot2)

expr_long <- expr_for_boxplot %>%
  pivot_longer(-c(Sample, Dataset), names_to = "Gene", values_to = "Expression")

expr_long$Sample <- factor(expr_long$Sample, levels = expr_for_boxplot$Sample[order(expr_for_boxplot$Dataset)])

# Compute vertical boundaries and label positions
sample_counts <- table(expr_for_boxplot$Dataset)
dataset_boundaries <- cumsum(as.numeric(sample_counts))
dataset_midpoints <- cumsum(c(0, as.numeric(sample_counts[-length(sample_counts)]))) + as.numeric(sample_counts) / 2
dataset_names <- names(sample_counts)

# Plot
ggplot(expr_long, aes(x = Sample, y = Expression, fill = Dataset)) +
  geom_boxplot(outlier.size = 0.2, outlier.alpha = 0.2, width = 0.4) +
  geom_vline(xintercept = dataset_boundaries[-length(dataset_boundaries)] + 0.5,
             linetype = "dashed", color = "grey40") +
  annotate("text", x = dataset_midpoints,
           y = max(expr_long$Expression, na.rm = TRUE) + 0.5,
           label = dataset_names,
           angle = 45, hjust = 0, size = 1.5) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(title = "Expression Distribution per Sample (Post-ComBat, Including Test Set)",
       x = "Samples (Grouped by Dataset)", y = "Expression") +
  scale_fill_brewer(palette = "Set2")
```

```{r}
library(glmnet)

train_matrix <- as.matrix(combat_train_t_aligned)
test_matrix <- as.matrix(combat_test_t_aligned)

train_y_bin <- ifelse(train_labels == "Asthma", 1, 0)

lasso_model <- cv.glmnet(train_matrix, train_y_bin, family = "binomial", alpha = 1)
lasso_probs <- predict(lasso_model, newx = test_matrix, s = "lambda.min", type = "response")
lasso_preds <- factor(ifelse(lasso_probs > 0.5, "Asthma", "Control"))

confusionMatrix(lasso_preds, test_labels)

```

```{r}
library(limma)
group_labels <- factor(ifelse(grepl("TRUE", pheno_data_40732$characteristics_ch1), "Asthma", "Control"))
design <- model.matrix(~0 + group_labels)
colnames(design) <- levels(group_labels)

fit <- lmFit(expr_matrix, design)
contrast <- makeContrasts(Asthma - Control, levels = design)
fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)

de_results <- topTable(fit2, number = Inf, sort.by = "P")  # This will now match your matrix


# 6. View top hits
head(de_results)
```

```{r}
library(randomForest)

# Transpose gene expression: rows = samples, columns = genes
expr_t <- as.data.frame(t(expr_matrix))

# Select top DE genes
top_de_genes <- rownames(de_results)[de_results$adj.P.Val < 0.05]
top_de_genes <- head(top_de_genes, 500)

# Select top variable genes
gene_vars <- apply(expr_t, 2, var, na.rm = TRUE)
top_var_genes <- names(sort(gene_vars, decreasing = TRUE))[1:500]

# Overlap
common_genes <- intersect(top_de_genes, top_var_genes)

# ---- CLEAN SUBSET FUNCTION ----
clean_subset <- function(df, genes) {
  # Force column names to be characters
  colnames(df) <- as.character(colnames(df))
  
  # Ensure input gene list is also character
  genes <- as.character(unlist(genes))  # If genes is a list, flatten it safely
  
  genes_present <- intersect(genes, colnames(df))
  
  if (length(genes_present) == 0) stop("No matching genes found in the data.")
  
  subset <- df[, genes_present, drop = FALSE]
  
  # Keep only columns that are fully numeric and finite
  keep_cols <- sapply(subset, function(x) is.numeric(x) && all(is.finite(x)))
  
  subset <- subset[, keep_cols, drop = FALSE]
  
  as.data.frame(subset)
}


# Clean feature sets
X_de     <- clean_subset(expr_t, top_de_genes)
X_var    <- clean_subset(expr_t, top_var_genes)
X_common <- clean_subset(expr_t, common_genes)

# Train random forest models
set.seed(3888)
rf_de     <- randomForest(x = X_de, y = labels)
rf_var    <- randomForest(x = X_var, y = labels)
rf_common <- randomForest(x = X_common, y = labels)

# Print training accuracy
cat("Accuracy (DE):",      mean(rf_de$predicted     == labels), "\n")
cat("Accuracy (Var):",     mean(rf_var$predicted    == labels), "\n")
cat("Accuracy (DE ∩ Var):", mean(rf_common$predicted == labels), "\n")

```

```{r}
# Check what's actually in your expression matrix
head(rownames(expr_matrix), 10)

# Check DE gene names
head(top_de_genes, 10)

# After transpose:
head(colnames(expr_t), 10)

```

```{r}
pca_6 = prcomp(t(expr_matrix_67472))
batch_gse67472 = ifelse(pca_6$x[,"PC1"] > 0, "batch_1", "batch_2")
pca_6$x %>%
  data.frame() %>%
  mutate(labels = labels_67472) %>%
  ggplot(aes(x = PC1, y = PC2, color = labels)) + geom_point() + theme_bw()

table(batch_gse67472, labels_67472)
```

```{r}
library(caret)
library(randomForest)

datasets <- list(
  GSE40732 = expr_matrix_40732,
  GSE67472 = expr_matrix_67472,
  GSE143303 = expr_matrix_gse143303_cleaned,
  GSE40888 = expr_matrix_40888_clean,
  GSE64913 = expr_matrix_64913_clean
)


# Set up 5-fold cross-validation
set.seed(3888)
cv_folds <- createFolds(train_labels, k = 5, returnTrain = TRUE)

# Placeholder for accuracy results
cv_accuracies <- c()

# Run CV
for (i in seq_along(cv_folds)) {
  cat("Fold", i, "\n")
  
  # Split data
  train_idx <- cv_folds[[i]]
  test_idx  <- setdiff(seq_along(train_labels), train_idx)
  
  X_train <- combat_train_t_aligned[train_idx, ]
  y_train <- train_labels[train_idx]
  
  X_test  <- combat_train_t_aligned[test_idx, ]
  y_test  <- train_labels[test_idx]
  
  # Train model
  rf_model <- randomForest(x = X_train, y = y_train, ntree = 500)
  
  # Predict
  preds <- predict(rf_model, newdata = X_test)
  
  # Accuracy
  acc <- mean(preds == y_test)
  cv_accuracies[i] <- acc
  cat("  Accuracy:", round(acc, 3), "\n")
}

# Final CV score
cat("\nAverage 5-Fold CV Accuracy:", round(mean(cv_accuracies), 3), "\n")

```

```{r}
# Load necessary library
library(ggplot2)

# Transpose the matrix: rows = samples, columns = genes
expr_40732_t <- t(expr_matrix_40732)

# Perform PCA
pca_40732 <- prcomp(expr_40732_t, scale. = TRUE, center = TRUE)

# Create PCA data frame
pca_df_40732 <- data.frame(
  PC1 = pca_40732$x[, 1],
  PC2 = pca_40732$x[, 2],
  Sample = rownames(pca_40732$x),
  Label = ifelse(grepl("TRUE", pheno_data_40732$characteristics_ch1), "Asthma", "Control")
)

# Plot PCA
ggplot(pca_df_40732, aes(x = PC1, y = PC2, color = Label)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "PCA of GSE40732 Samples",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_manual(values = c("Asthma" = "#d95f02", "Control" = "#1b9e77"))

```

```{r}
# Load required library
library(ggplot2)

# Step 1: Prepare the uncorrected merged expression matrix
# (Assumes 'merged_full' was created earlier and includes all datasets)
# merged_full should be a matrix with genes as rows and samples as columns

# Step 2: Transpose so samples are rows, genes are columns
expr_precombat <- t(merged_full)

# Step 3: Perform PCA
pca_precombat <- prcomp(expr_precombat, scale. = TRUE, center = TRUE)

# Step 4: Create a data frame for plotting
pca_df_precombat <- data.frame(PC1 = pca_precombat$x[, 1],
                               PC2 = pca_precombat$x[, 2],
                               Dataset = full_batch_labels)  # Already matches column order of merged_full

# Step 5: Plot the PCA
ggplot(pca_df_precombat, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "PCA of All Samples (Pre-ComBat)",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_brewer(palette = "Set2")

```

```{r}
# Transpose the corrected matrix so that rows = samples and columns = genes
combat_pca_data <- t(combat_corrected_full)

# Perform PCA on the transposed (sample-wise) matrix
pca_result <- prcomp(combat_pca_data, scale. = TRUE, center = TRUE)

# Build the PCA data frame
pca_df <- data.frame(PC1 = pca_result$x[, 1],
                     PC2 = pca_result$x[, 2],
                     Dataset = full_batch_labels)  # Match original batch/dataset labels

# Plot the PCA
library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "PCA of All Samples (Post-ComBat)",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_brewer(palette = "Set2") 

```

```{r}
library(caret)
library(randomForest)
library(sva)  # for ComBat

raw_expr_matrix <- cbind(
  expr_matrix_64913_clean[common_genes_all, ],
  expr_matrix_40888_clean[common_genes_all, ],
  expr_matrix_gse143303_cleaned[common_genes_all, ],
  expr_matrix_67472[common_genes_all, ]
)


set.seed(3888)
cv_folds <- createFolds(train_labels, k = 5, returnTrain = TRUE)
cv_accuracies <- c()

for (i in seq_along(cv_folds)) {
  cat("Fold", i, "\n")
  
  # Split data
  train_idx <- cv_folds[[i]]
  test_idx  <- setdiff(seq_along(train_labels), train_idx)
  
  expr_train <- t(raw_expr_matrix[, train_idx])
  expr_test  <- t(raw_expr_matrix[, test_idx])
  
  y_train <- train_labels[train_idx]
  y_test  <- train_labels[test_idx]
  
  # Batch labels for ComBat (assumed known)
  batch_train <- batch_labels[train_idx]
  batch_test  <- batch_labels[test_idx]
  
  # Apply ComBat on training data only
  combat_train <- ComBat(dat = t(expr_train), batch = batch_train, par.prior = TRUE)
  
  # Estimate ComBat params from training data
  combat_model <- ComBat(dat = t(expr_train), batch = batch_train, mod = NULL, par.prior = TRUE, prior.plots = FALSE)
  
  # Apply the same parameters to test data
  combat_test <- ComBat(dat = t(expr_test), batch = batch_test, mod = NULL, par.prior = TRUE, prior.plots = FALSE)
  
  # Transpose back for model
  X_train <- t(combat_train)
  X_test  <- t(combat_test)
  
  # Train model
  rf_model <- randomForest(x = X_train, y = y_train, ntree = 500)
  
  # Predict
  preds <- predict(rf_model, newdata = X_test)
  
  # Accuracy
  acc <- mean(preds == y_test)
  cv_accuracies[i] <- acc
  cat("  Accuracy:", round(acc, 3), "\n")
}

cat("\nAverage 5-Fold CV Accuracy (no leakage):", round(mean(cv_accuracies), 3), "\n")
```

```{r}
# Perform PCA on GSE67472 expression data (samples should be rows)
expr_67472_t <- t(expr_matrix_67472)
pca_67472 <- prcomp(expr_67472_t, center = TRUE, scale. = TRUE)

# Option 1: Split based on PC1 threshold
# Check distribution of PC1
plot(pca_67472$x[,1], main = "PC1 values for GSE67472", ylab = "PC1", xlab = "Sample Index")

# Method: split by sign of PC1 (or set a custom threshold if needed)
split_labels <- ifelse(pca_67472$x[, 1] > 0, "GSE67472_batch1", "GSE67472_batch2")

# Check distribution
table(split_labels)

# Create two separate expression matrices
samples_batch1 <- rownames(pca_67472$x)[split_labels == "GSE67472_batch1"]
samples_batch2 <- rownames(pca_67472$x)[split_labels == "GSE67472_batch2"]

expr_matrix_67472_batch1 <- expr_matrix_67472[, samples_batch1]
expr_matrix_67472_batch2 <- expr_matrix_67472[, samples_batch2]

# Optional: verify
dim(expr_matrix_67472_batch1)
dim(expr_matrix_67472_batch2)

```

```{r}
library(sva)
library(ggplot2)

# Step 1: Find common genes across ALL datasets
common_genes_all <- Reduce(intersect, list(
  rownames(expr_matrix_64913_clean),
  rownames(expr_matrix_40888_clean),
  rownames(expr_matrix_gse143303_cleaned),
  rownames(expr_matrix_40732),
  rownames(expr_matrix_67472_batch1),
  rownames(expr_matrix_67472_batch2)
))

# Step 2: Combine all datasets using only common genes
merged_full <- cbind(
  expr_matrix_64913_clean[common_genes_all, ],
  expr_matrix_40888_clean[common_genes_all, ],
  expr_matrix_gse143303_cleaned[common_genes_all, ],
  expr_matrix_67472_batch1[common_genes_all, ],
  expr_matrix_67472_batch2[common_genes_all, ],
  expr_matrix_40732[common_genes_all, ]
)

# Step 3: Create batch labels (must match column order)
full_batch_labels <- c(
  rep("GSE64913", ncol(expr_matrix_64913_clean)),
  rep("GSE40888", ncol(expr_matrix_40888_clean)),
  rep("GSE143303", ncol(expr_matrix_gse143303_cleaned)),
  rep("GSE67472_batch1", ncol(expr_matrix_67472_batch1)),
  rep("GSE67472_batch2", ncol(expr_matrix_67472_batch2)),
  rep("GSE40732", ncol(expr_matrix_40732))
)

# Step 4: Apply ComBat batch correction
combat_corrected <- ComBat(dat = merged_full, batch = full_batch_labels, par.prior = TRUE)

# Step 5: PCA on batch-corrected data (samples as rows)
combat_pca_data <- t(combat_corrected)
pca_res <- prcomp(combat_pca_data, scale. = TRUE, center = TRUE)

# Step 6: Create PCA dataframe
pca_df <- data.frame(PC1 = pca_res$x[,1],
                     PC2 = pca_res$x[,2],
                     Dataset = full_batch_labels)

# Step 7: Plot PCA
ggplot(pca_df, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "PCA of All Samples (Post-ComBat, GSE67472 split)",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_brewer(palette = "Set2")

```

```{r}
library(ggplot2)

# Transpose expression matrix: samples as rows, genes as columns
expr_40888_t <- t(expr_matrix_40888_clean)

# Perform PCA
pca_40888 <- prcomp(expr_40888_t, center = TRUE, scale. = TRUE)

# Create PCA data frame
pca_df_40888 <- data.frame(
  PC1 = pca_40888$x[, 1],
  PC2 = pca_40888$x[, 2],
  Sample = rownames(pca_40888$x),
  Label = ifelse(grepl("asthmatic", pheno_data_40888$`characteristics_ch1.3`, ignore.case = TRUE),
                 "Asthma", "Control")
)

# Plot PCA
ggplot(pca_df_40888, aes(x = PC1, y = PC2, color = Label)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "PCA of GSE40888 Samples",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_manual(values = c("Asthma" = "#d95f02", "Control" = "#1b9e77"))

```

```{r}
library(ggplot2)
library(sva)

# Perform PCA again
expr_40888_t <- t(expr_matrix_40888_clean)
pca_40888 <- prcomp(expr_40888_t, center = TRUE, scale. = TRUE)

# Plot PC1 to visually check gap if you want
plot(pca_40888$x[,1], main = "PC1 values for GSE40888", ylab = "PC1", xlab = "Sample Index")

# Split by PC1 (this seems to be your real separation axis)
split_labels_40888 <- ifelse(pca_40888$x[,1] > 0, "GSE40888_batch1", "GSE40888_batch2")

table(split_labels_40888)  # check counts

# Create two matrices
samples_batch1_40888 <- rownames(pca_40888$x)[split_labels_40888 == "GSE40888_batch1"]
samples_batch2_40888 <- rownames(pca_40888$x)[split_labels_40888 == "GSE40888_batch2"]

expr_matrix_40888_batch1 <- expr_matrix_40888_clean[, samples_batch1_40888]
expr_matrix_40888_batch2 <- expr_matrix_40888_clean[, samples_batch2_40888]


# ------------------------------
# Now do PCA of all datasets incl. splits
# ------------------------------

# Step 4: Get common genes across ALL datasets including splits
common_genes_all <- Reduce(intersect, list(
  rownames(expr_matrix_64913_clean),
  rownames(expr_matrix_gse143303_cleaned),
  rownames(expr_matrix_40732),
  rownames(expr_matrix_67472_batch1),
  rownames(expr_matrix_67472_batch2),
  rownames(expr_matrix_40888_batch1),
  rownames(expr_matrix_40888_batch2)
))

# Step 5: Merge all datasets
merged_full <- cbind(
  expr_matrix_64913_clean[common_genes_all, ],
  expr_matrix_40888_batch1[common_genes_all, ],
  expr_matrix_40888_batch2[common_genes_all, ],
  expr_matrix_gse143303_cleaned[common_genes_all, ],
  expr_matrix_67472_batch1[common_genes_all, ],
  expr_matrix_67472_batch2[common_genes_all, ],
  expr_matrix_40732[common_genes_all, ]
)

# Step 6: Create batch labels
full_batch_labels <- c(
  rep("GSE64913", ncol(expr_matrix_64913_clean)),
  rep("GSE40888_batch1", ncol(expr_matrix_40888_batch1)),
  rep("GSE40888_batch2", ncol(expr_matrix_40888_batch2)),
  rep("GSE143303", ncol(expr_matrix_gse143303_cleaned)),
  rep("GSE67472_batch1", ncol(expr_matrix_67472_batch1)),
  rep("GSE67472_batch2", ncol(expr_matrix_67472_batch2)),
  rep("GSE40732", ncol(expr_matrix_40732))
)

# Step 7: Apply ComBat batch correction
combat_corrected <- ComBat(dat = merged_full, batch = full_batch_labels, par.prior = TRUE)

write.csv(combat_corrected, "expression_7datasets_postCombat.csv")

# Step 8: PCA on batch-corrected data
combat_pca_data <- t(combat_corrected)
pca_res <- prcomp(combat_pca_data, scale. = TRUE, center = TRUE)

saveRDS(combat_corrected, file = "frozen_combat_corrected_expr_primary_model.rds")

# Step 9: Plot PCA
pca_df <- data.frame(PC1 = pca_res$x[,1],
                     PC2 = pca_res$x[,2],
                     Dataset = full_batch_labels)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Dataset)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "PCA of All Samples (Post-ComBat, GSE67472 + GSE40888 split)",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  scale_color_brewer(palette = "Set2")

```

```{r}
library(caret)
library(dplyr)
library(ggplot2)

# Example expression data
# expr_matrix: samples x genes
# labels: factor of class labels (same order as expr_matrix rows)

# Replace these with your actual data
# Example dummy data for illustration:
set.seed(3888)
n_samples <- 100
n_genes <- 5  # adjust for Top2, Top5, Top8 etc.
expr_matrix <- matrix(rnorm(n_samples * n_genes), ncol = n_genes)
colnames(expr_matrix) <- paste0("Gene", 1:n_genes)
labels <- factor(sample(c("Asthma", "Control"), n_samples, replace = TRUE))

# Data frame for caret
data_df <- as.data.frame(expr_matrix)
data_df$Class <- labels

# Define models to run
model_list <- c("rf", "svmLinear", "knn", "nb", "rpart")  # Random Forest, SVM, KNN, Naive Bayes, Decision Tree

# Create trainControl object
ctrl <- trainControl(method = "cv", number = 5, savePredictions = "final")

# Store results
results <- list()

# Run models
for (model_name in model_list) {
  set.seed(3888)
  cat("Training model:", model_name, "\n")
  
  fit <- train(Class ~ ., data = data_df,
               method = model_name,
               trControl = ctrl,
               preProcess = c("center", "scale"))
  
  results[[model_name]] <- fit
}

```

```{r}
# Extract accuracy results
accuracy_df <- bind_rows(lapply(names(results), function(name) {
  fit <- results[[name]]
  data.frame(
    Model = name,
    Accuracy = fit$resample$Accuracy
  )
}))

# Optional: change model names to match your plot labels
accuracy_df$Model <- factor(accuracy_df$Model, 
                            levels = c("rf", "svmLinear", "knn", "nb", "rpart"),
                            labels = c("RF", "SVM", "KNN", "NB", "DT"))

head(accuracy_df)

```

```{r}
ggplot(accuracy_df, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "white", outlier.color = "black", width = 0.6) +
  scale_fill_manual(values = c("RF" = "red", "SVM" = "green", "KNN" = "blue", 
                               "NB" = "purple", "DT" = "orange")) +
  theme_minimal(base_size = 14) +
  labs(title = "Comparison of Model Accuracies",
       y = "Cross-Validated Accuracy",
       x = "Model") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0.4, 0.8))

```

```{r}
library(caret)
library(dplyr)
library(ggplot2)
library(limma)

# ---- INPUT DATA: replace these with your real data ----
# expr_matrix: genes x samples matrix
# labels: factor of sample labels (same order as columns of expr_matrix)

# Example mock data for testing (replace with yours!)
set.seed(3888)
expr_matrix <- matrix(rnorm(1000 * 80), nrow = 1000)  # 1000 genes x 80 samples
rownames(expr_matrix) <- paste0("Gene", 1:1000)
labels <- factor(rep(c("Asthma", "Control"), each = 40))

# ---- FUNCTION to get top N DE genes ----
get_top_genes <- function(expr_matrix, labels, top_n) {
  design <- model.matrix(~labels)
  fit <- lmFit(expr_matrix, design)
  fit <- eBayes(fit)
  topTableRes <- topTable(fit, number = top_n, sort.by = "P")
  selected_genes <- rownames(topTableRes)
  return(selected_genes)
}

# ---- RUN CLASSIFIERS FOR DIFFERENT TOP GENE SETS ----
results_list <- list()
top_gene_sets <- c(2, 5, 8)

for (n_genes in top_gene_sets) {
  cat("Running models for top", n_genes, "DE genes...\n")
  
  # Step 1: select top N genes
  top_genes <- get_top_genes(expr_matrix, labels, n_genes)
  
  # Step 2: prepare data (transpose to samples x genes)
  data_df <- as.data.frame(t(expr_matrix[top_genes, ]))
  data_df$Class <- labels
  
  # Step 3: set up trainControl
  ctrl <- trainControl(method = "cv", number = 5, savePredictions = "final")
  
  # Step 4: run models
  model_list <- c("rf", "svmLinear", "knn", "nb", "rpart")
  
  for (model_name in model_list) {
    set.seed(3888)
    fit <- train(Class ~ ., data = data_df,
                 method = model_name,
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
    
    # Store results
    model_results <- data.frame(
      Model = model_name,
      Accuracy = fit$resample$Accuracy,
      TopGenes = paste0("Top", n_genes, " DE Genes")
    )
    
    results_list[[length(results_list) + 1]] <- model_results
  }
}

# ---- COMBINE RESULTS ----
results_df <- bind_rows(results_list)

# Clean up model names for plotting
results_df$Model <- factor(results_df$Model, 
                           levels = c("rf", "svmLinear", "knn", "nb", "rpart"),
                           labels = c("RF", "SVM", "KNN", "NB", "DT"))

# ---- PLOT ----
ggplot(results_df, aes(x = Model, y = Accuracy, fill = TopGenes)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "white", outlier.color = "black", width = 0.6) +
  scale_fill_manual(values = c("Top2 DE Genes" = "red",
                               "Top5 DE Genes" = "green",
                               "Top8 DE Genes" = "blue")) +
  theme_minimal(base_size = 14) +
  labs(title = "Comparison of Model Accuracies across Different Top DE Gene Sets",
       y = "Cross-Validated Accuracy",
       x = "Model") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0.4, 0.8))

```

```{r}
library(caret)
library(dplyr)
library(ggplot2)
library(limma)



# ---- SET YOUR ACTUAL DATA ----
# Replace combat_train_t_aligned and train_labels with your objects
expr_matrix <- t(combat_train_t_aligned)  # genes x samples (transpose!)
labels <- train_labels                    # factor labels (same order as columns of expr_matrix)

# ---- FUNCTION to get top N DE genes ----
get_top_genes <- function(expr_matrix, labels, top_n) {
  design <- model.matrix(~labels)
  fit <- lmFit(expr_matrix, design)
  fit <- eBayes(fit)
  topTableRes <- topTable(fit, number = top_n, sort.by = "P")
  selected_genes <- rownames(topTableRes)
  return(selected_genes)
}

# ---- RUN CLASSIFIERS FOR DIFFERENT TOP GENE SETS ----
results_list <- list()
top_gene_sets <- c(2, 5, 8)

for (n_genes in top_gene_sets) {
  cat("Running models for top", n_genes, "DE genes...\n")
  
  # Step 1: select top N genes
  top_genes <- get_top_genes(expr_matrix, labels, n_genes)
  
  # Step 2: prepare data for caret (samples x genes)
  data_df <- as.data.frame(t(expr_matrix[top_genes, ]))  # transpose back
  data_df$Class <- labels
  
  # Step 3: set up trainControl
  ctrl <- trainControl(method = "cv", number = 5, savePredictions = "final")
  
  # Step 4: run models
  model_list <- c("rf", "svmLinear", "knn", "nb", "rpart")
  
  for (model_name in model_list) {
    set.seed(3888)
    fit <- train(Class ~ ., data = data_df,
                 method = model_name,
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
    
    # Store results
    model_results <- data.frame(
      Model = model_name,
      Accuracy = fit$resample$Accuracy,
      TopGenes = paste0("Top", n_genes, " DE Genes")
    )
    
    results_list[[length(results_list) + 1]] <- model_results
  }
}

# ---- COMBINE RESULTS ----
results_df <- bind_rows(results_list)

# Clean up model names for plotting
results_df$Model <- factor(results_df$Model, 
                           levels = c("rf", "svmLinear", "knn", "nb", "rpart"),
                           labels = c("RF", "SVM", "KNN", "NB", "DT"))

# ---- PLOT ----
ggplot(results_df, aes(x = Model, y = Accuracy, fill = TopGenes)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "white", outlier.color = "black", width = 0.6) +
  scale_fill_manual(values = c("Top2 DE Genes" = "red",
                               "Top5 DE Genes" = "green",
                               "Top8 DE Genes" = "blue")) +
  theme_minimal(base_size = 14) +
  labs(title = "Comparison of Model Accuracies across Different Top DE Gene Sets",
       y = "Cross-Validated Accuracy",
       x = "Model") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0.4, 0.8))

```

```{r}
design <- model.matrix(~labels)
fit <- lmFit(expr_matrix, design)
fit <- eBayes(fit)
top_20_table <- topTable(fit, number = 20, sort.by = "P")
print(top_20_table)

# Get the list of top 20 gene names
top_20_genes <- rownames(top_20_table)

# Save as .rds file
saveRDS(top_20_genes, file = "top_20_genes.rds")
```

```{r}
library(caret)
library(dplyr)
library(ggplot2)
library(limma)

# ---- Expression matrix and labels ----
expr_matrix <- t(combat_train_t_aligned)  # genes x samples
labels <- train_labels                    # factor vector

# ---- Function to get top N DE genes ----
get_top_genes <- function(expr_matrix, labels, top_n) {
  design <- model.matrix(~labels)
  fit <- lmFit(expr_matrix, design)
  fit <- eBayes(fit)
  topTableRes <- topTable(fit, number = top_n, sort.by = "P")
  selected_genes <- rownames(topTableRes)
  return(selected_genes)
}

# ---- Model training ----
results_list <- list()
top_gene_sets <- c(8, 10, 15, 20, 25)

for (n_genes in top_gene_sets) {
  cat("Running models for top", n_genes, "DE genes...\n")
  
  # Step 1: Get top genes
  top_genes <- get_top_genes(expr_matrix, labels, n_genes)
  
  # Step 2: Prepare data for caret
  data_df <- as.data.frame(t(expr_matrix[top_genes, ]))  # samples x genes
  data_df$Class <- labels
  
  # Step 3: Set up cross-validation
  ctrl <- trainControl(method = "cv", number = 5, savePredictions = "final")
  
  # Step 4: Train RF and KNN only
  for (model_name in c("rf", "knn")) {
    set.seed(3888)
    fit <- train(Class ~ ., data = data_df,
                 method = model_name,
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
    
    # Save the RF model with top 20 genes
    if (n_genes == 20 && model_name == "rf") {
      saveRDS(fit, file = "rf_model_top20_genes.rds")
      cat("Saved RF model trained on top 20 DE genes to 'rf_model_top20_genes.rds'\n")
    }
    
    # Store results
    model_results <- data.frame(
      Model = model_name,
      Accuracy = fit$resample$Accuracy,
      TopGenes = paste0("Top", n_genes, " DE Genes")
    )
    
    results_list[[length(results_list) + 1]] <- model_results
  }
}

# ---- Combine and Plot ----
results_df <- bind_rows(results_list)

results_df$Model <- factor(results_df$Model,
                           levels = c("rf", "knn"),
                           labels = c("RF", "KNN"))

ggplot(results_df, aes(x = Model, y = Accuracy, fill = TopGenes)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "white", outlier.color = "black", width = 0.6) +
  scale_fill_manual(values = c("Top8 DE Genes" = "blue",
                               "Top10 DE Genes" = "purple",
                               "Top15 DE Genes" = "orange",
                               "Top20 DE Genes" = "darkgreen",
                               "Top25 DE Genes" = "brown")) +
  theme_minimal(base_size = 14) +
  labs(title = "RF and KNN Accuracy with Increasing DE Gene Sets",
       y = "Cross-Validated Accuracy",
       x = "Model") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0.4, 0.9))

```

```{r}
library(caret)
library(dplyr)
library(ggplot2)
library(limma)

# ---- Expression matrix and labels ----
expr_matrix <- t(combat_train_t_aligned)  # genes x samples
labels <- train_labels                    # factor labels

# ---- Function to get top N DE genes ----
get_top_genes <- function(expr_matrix, labels, top_n) {
  design <- model.matrix(~labels)
  fit <- lmFit(expr_matrix, design)
  fit <- eBayes(fit)
  topTableRes <- topTable(fit, number = top_n, sort.by = "P")
  selected_genes <- rownames(topTableRes)
  return(selected_genes)
}

# ---- Run models for selected top gene counts ----
results_list <- list()
top_gene_sets <- c(10, 15, 20, 25)

for (n_genes in top_gene_sets) {
  cat("Running models for top", n_genes, "DE genes...\n")
  
  # Select top N DE genes
  top_genes <- get_top_genes(expr_matrix, labels, n_genes)
  
  # Prepare data
  data_df <- as.data.frame(t(expr_matrix[top_genes, ]))
  data_df$Class <- labels
  
  # Cross-validation setup
  ctrl <- trainControl(method = "cv", number = 5, savePredictions = "final")
  
  # Models to train
  model_list <- c("rf", "svmLinear", "knn", "nb", "rpart")
  
  for (model_name in model_list) {
    set.seed(3888)
    fit <- train(Class ~ ., data = data_df,
                 method = model_name,
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
    
    # Store results
    model_results <- data.frame(
      Model = model_name,
      Accuracy = fit$resample$Accuracy,
      TopGenes = paste0("Top", n_genes, " DE Genes")
    )
    
    results_list[[length(results_list) + 1]] <- model_results
  }
}

# ---- Combine and Plot ----
results_df <- bind_rows(results_list)

# Clean up model names
results_df$Model <- factor(results_df$Model,
                           levels = c("rf", "svmLinear", "knn", "nb", "rpart"),
                           labels = c("RF", "SVM", "KNN", "NB", "DT"))

# Plotting
ggplot(results_df, aes(x = Model, y = Accuracy, fill = TopGenes)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "white", outlier.color = "black", width = 0.6) +
  scale_fill_manual(values = c("Top10 DE Genes" = "purple",
                               "Top15 DE Genes" = "orange",
                               "Top20 DE Genes" = "darkgreen",
                               "Top25 DE Genes" = "brown")) +
  theme_minimal(base_size = 14) +
  labs(title = "Model Accuracy with Top 10–25 DE Genes",
       y = "Cross-Validated Accuracy",
       x = "Model") +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0.4, 0.9))

```

```{r}

```

```{r}
library(caret)
library(limma)
library(dplyr)
library(ggplot2)

# These should be BEFORE you applied ComBat anywhere
raw_67472 <- expr_matrix_67472
raw_40888 <- expr_matrix_40888_clean


# ---- STEP 1: Define all labels ----
labels_64913 <- ifelse(grepl("Asthmatic", pheno_data_64913$characteristics_ch1.1, ignore.case = TRUE),
                       "Asthma", "Control")

labels_40888_all <- ifelse(grepl("asthmatic", pheno_data_40888$characteristics_ch1.3, ignore.case = TRUE),
                           "Asthma", "Control")
labels_40888 <- labels_40888_all
names(labels_40888) <- rownames(pheno_data_40888)

labels_143303 <- ifelse(pheno_data_gse143303$characteristics_ch1.1 == "disease status: Severe asthma",
                        "Asthma", "Control")
names(labels_143303) <- rownames(pheno_data_gse143303)

labels_67472_all <- ifelse(grepl("asthma", pheno_data_67472$`disease state:ch1`, ignore.case = TRUE),
                           "Asthma", "Control")
labels_67472 <- labels_67472_all
names(labels_67472) <- rownames(pheno_data_67472)

labels_40732 <- ifelse(grepl("TRUE", pheno_data_40732$characteristics_ch1), 
                       "Asthma", "Control")
names(labels_40732) <- rownames(pheno_data_40732)

expr_matrix_64913_clean <- data.matrix(expr_matrix_64913_clean)
expr_matrix_40888_batch1 <- data.matrix(expr_matrix_40888_batch1)
expr_matrix_40888_batch2 <- data.matrix(expr_matrix_40888_batch2)
expr_matrix_gse143303_cleaned <- data.matrix(expr_matrix_gse143303_cleaned)
expr_matrix_67472_batch1 <- data.matrix(expr_matrix_67472_batch1)
expr_matrix_67472_batch2 <- data.matrix(expr_matrix_67472_batch2)
expr_matrix_40732 <- data.matrix(expr_matrix_40732)

# ---- STEP 2: Combine datasets ----
combat_train_t_aligned <- t(cbind(
  expr_matrix_64913_clean[common_genes, ],
  expr_matrix_40888_batch1[common_genes, ],
  expr_matrix_40888_batch2[common_genes, ],
  expr_matrix_gse143303_cleaned[common_genes, ],
  expr_matrix_67472_batch1[common_genes, ],
  expr_matrix_67472_batch2[common_genes, ],
  expr_matrix_40732[common_genes, ]
))

# ---- STEP 3: Create batch labels ----
full_batch_labels <- c(
  rep("GSE64913", ncol(expr_matrix_64913_clean)),
  rep("GSE40888_batch1", ncol(expr_matrix_40888_batch1)),
  rep("GSE40888_batch2", ncol(expr_matrix_40888_batch2)),
  rep("GSE143303", ncol(expr_matrix_gse143303_cleaned)),
  rep("GSE67472_batch1", ncol(expr_matrix_67472_batch1)),
  rep("GSE67472_batch2", ncol(expr_matrix_67472_batch2)),
  rep("GSE40732", ncol(expr_matrix_40732))
)

# ---- STEP 4: Create labels ----
train_labels <- c(
  labels_64913,
  labels_40888[colnames(expr_matrix_40888_batch1)],
  labels_40888[colnames(expr_matrix_40888_batch2)],
  labels_143303,
  labels_67472[colnames(expr_matrix_67472_batch1)],
  labels_67472[colnames(expr_matrix_67472_batch2)],
  labels_40732
)

# ---- STEP 5: Expression matrix ----
expr_matrix <- combat_train_t_aligned
labels <- train_labels
groups <- full_batch_labels

# ---- STEP 6: DE gene selection ----
get_top_genes <- function(expr_matrix, labels, top_n) {
  expr_matrix <- data.matrix(expr_matrix)         # 🚨 guarantees numeric matrix
  design <- model.matrix(~labels)
  fit <- lmFit(t(expr_matrix), design)
  fit <- eBayes(fit)
  topTableRes <- topTable(fit, number = top_n, sort.by = "P")
  rownames(topTableRes)
}


# ---- STEP 7: 7-FOLD dataset CV ----
results_list <- list()
top_gene_sets <- c(2, 5, 8)
folds <- groupKFold(groups, k = length(unique(groups)))

for (n_genes in top_gene_sets) {
  cat("Running models for top", n_genes, "genes...\n")
  
  top_genes <- get_top_genes(expr_matrix, labels, n_genes)
  data_df <- as.data.frame(expr_matrix[, top_genes])
  data_df$Class <- labels
  
  ctrl <- trainControl(index = folds,
                       savePredictions = "final")
  
  model_list <- c("rf", "svmLinear", "knn", "nb", "rpart")
  
  for (model_name in model_list) {
    set.seed(3888)
    fit <- train(Class ~ ., data = data_df,
                 method = model_name,
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
    
    model_results <- data.frame(
      Model = model_name,
      Accuracy = fit$resample$Accuracy,
      TopGenes = paste0("Top", n_genes)
    )
    results_list[[length(results_list) + 1]] <- model_results
  }
}

# ---- STEP 8: Plot results ----
results_df <- bind_rows(results_list)

results_df$Model <- factor(results_df$Model,
                           levels = c("rf", "svmLinear", "knn", "nb", "rpart"),
                           labels = c("RF", "SVM", "KNN", "NB", "DT"))

ggplot(results_df, aes(x = Model, y = Accuracy, fill = TopGenes)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "white", outlier.color = "black") +
  theme_minimal(base_size = 14) +
  labs(title = "7 Dataset Leave-One-Dataset-Out CV Accuracy",
       y = "Accuracy", x = "Model") +
  scale_fill_manual(values = c("Top2" = "red", "Top5" = "green", "Top8" = "blue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
library(pROC)

# ---- EXTRACT ROC CURVES FROM EXISTING MODELS ----
# We assume you have kept your trained models in 'results'

# (OPTIONAL) If you want to reuse 'results_list' from your training code:
# You would have to also save the model fits.
# Your current code only stores accuracies, not the full models.

# BUT if you had something like this in training:
# results_models[[paste0("Top", n_genes, "_", model_name)]] <- fit
# Then you could use that.

# So here's an example assuming you saved full models like this:
# (This would be your additional training block modification)

results_models <- list()

for (n_genes in top_gene_sets) {
  cat("Running models for ROC extraction for top", n_genes, "DE genes...\n")
  
  # Select top N genes
  top_genes <- get_top_genes(expr_matrix, labels, n_genes)
  
  # Prepare data
  data_df <- as.data.frame(t(expr_matrix[top_genes, ]))
  data_df$Class <- labels
  
  # TrainControl for ROC
  ctrl <- trainControl(method = "cv", number = 5,
                       classProbs = TRUE, 
                       summaryFunction = twoClassSummary,
                       savePredictions = "final")
  
  model_list <- c("rf", "svmLinear", "knn", "nb", "rpart")
  
  for (model_name in model_list) {
    set.seed(3888)
    fit <- train(Class ~ ., data = data_df,
                 method = model_name,
                 trControl = ctrl,
                 preProcess = c("center", "scale"),
                 metric = "ROC")
    
    # Save full model
    results_models[[paste0("Top", n_genes, "_", model_name)]] <- fit
  }
}

# ---- NOW EXTRACT ROC CURVES ----

roc_list <- list()

for (name in names(results_models)) {
  fit <- results_models[[name]]
  preds <- fit$pred
  
  # Sometimes multiple resamples: choose best tuning parameter if needed
  if ("mtry" %in% names(preds)) {
    best_mtry <- fit$bestTune$mtry
    preds <- preds[preds$mtry == best_mtry, ]
  }
  if ("C" %in% names(preds)) { # for SVM
    best_C <- fit$bestTune$C
    preds <- preds[preds$C == best_C, ]
  }
  if ("k" %in% names(preds)) { # for kNN
    best_k <- fit$bestTune$k
    preds <- preds[preds$k == best_k, ]
  }
  
  # Compute ROC
  roc_obj <- roc(preds$obs, preds$Asthma, levels = c("Control", "Asthma"))
  roc_list[[name]] <- roc_obj
}

# ---- PLOT ROC CURVES ----
ggroc(roc_list, legacy.axes = TRUE) +
  ggtitle("ROC Curves for All Models (5-Fold CV)") +
  theme_minimal() +
  labs(x = "False Positive Rate", y = "True Positive Rate")


```

```{r}
rf_fit <- readRDS("rf_model_top20_genes.rds")

# Extract predictions from cross-validation
preds <- rf_fit$pred

# If tuning (e.g., mtry) was used, select predictions from the best hyperparameter
if ("mtry" %in% names(preds)) {
  preds <- preds[preds$mtry == rf_fit$bestTune$mtry, ]
}

```

```{r}
library(caret)

# Create confusion matrix
conf_mat <- confusionMatrix(preds$pred, preds$obs, positive = "Asthma")
print(conf_mat)

```

```{r}
library(pROC)

# Use predicted probabilities for ROC
roc_obj <- roc(preds$obs, preds$Asthma, levels = c("Control", "Asthma"))

# Plot ROC curve
plot(roc_obj, col = "blue", lwd = 2, main = "ROC Curve for RF (Top 20 DE Genes)")
auc_val <- auc(roc_obj)
cat("AUC:", round(auc_val, 3), "\n")

```

# Model 2
NOTE THAT THIS IS WHAT THE RIGHT EXPR MATRIXES ARE CALLED
datasets <- list(
  GSE40732 = expr_matrix_40732,
  GSE67472 = expr_matrix_67472,
  GSE143303 = expr_matrix_gse143303_cleaned,
  GSE40888 = expr_matrix_40888_clean,
  GSE64913 = expr_matrix_64913_clean
)

```{r}
length(common_genes_all)
```


```{r}
BiocManager::install("sva")
```


```{r}
# Load required packages
library(sva)
library(limma)
library(caret)

# ---- 1. Create Labels ----
# Dataset: GSE40732
labels_40732 <- ifelse(grepl("subtype: non", pheno_data_40732$characteristics_ch1.1), "healthy",
                 ifelse(grepl("subtype: reversible|subtype: PC20", pheno_data_40732$characteristics_ch1.1), "moderate", "severe"))

# Dataset: GSE67472
labels_67472 <- ifelse(pheno_data_67472$`disease state:ch1` == "asthma", "moderate", "healthy")

# Dataset: GSE143303
labels_143303 <- ifelse(grepl("Severe asthma", pheno_data_143303$characteristics_ch1.1), "severe", "healthy")

# Dataset: GSE64913
labels_64913 <- ifelse(grepl("Severe Asthmatic", pheno_data_64913$characteristics_ch1.1), "severe", "healthy")

# ---- 2. Combine Expression and Labels ----
common_genes_all <- Reduce(intersect, list(
  rownames(expr_matrix_64913_clean),
  rownames(expr_matrix_40732),
  rownames(expr_matrix_gse143303_cleaned),
  rownames(expr_matrix_67472)
))

# Expression matrix
raw_expr_matrix <- cbind(
  expr_matrix_64913_clean[common_genes_all, ],
  expr_matrix_40732[common_genes_all, ],
  expr_matrix_gse143303_cleaned[common_genes_all, ],
  expr_matrix_67472[common_genes_all, ]
)

# Labels
combined_labels <- c(labels_64913, labels_40732, labels_143303, labels_67472)
```

```{r}
pca_67472 <- prcomp(t(expr_matrix_67472[common_genes_all, ]), scale. = TRUE)
pc1_67472 <- pca_67472$x[, 1]

# Sub-batch labels
sub_batch_67472 <- ifelse(pc1_67472 >= 0, "batch1_67472", "batch2_67472")
```


```{r}
# Batch labels
batch_labels <- c(
  rep("GSE64913", ncol(expr_matrix_64913_clean)),
  rep("GSE40732", ncol(expr_matrix_40732)),
  rep("GSE143303", ncol(expr_matrix_gse143303_cleaned)),
  sub_batch_67472
)

# ---- 3. Apply ComBat with Mod Matrix ----
mod <- model.matrix(~ combined_labels)
combat_corrected_expr <- ComBat(dat = raw_expr_matrix, batch = batch_labels, mod = mod, par.prior = TRUE)

# ---- 4. Drop Healthy Samples ----
non_healthy_idx <- which(combined_labels != "healthy")
expr_non_healthy <- combat_corrected_expr[, non_healthy_idx]
labels_non_healthy <- combined_labels[non_healthy_idx]

# ---- 5. Differential Expression (Moderate vs Severe) ----
design <- model.matrix(~ 0 + factor(labels_non_healthy))
colnames(design) <- levels(factor(labels_non_healthy))
fit <- lmFit(expr_non_healthy, design)
contrast_matrix <- makeContrasts(severe - moderate, levels=design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
top_genes_df <- topTable(fit2, adjust="fdr", number=Inf)
```

```{r}
# pca on unmerged data (coloured by condition)
pca_unmerged <- prcomp(t(raw_expr_matrix), scale. = TRUE)
pca_df_unmerged <- data.frame(
  PC1 = pca_unmerged$x[,1],
  PC2 = pca_unmerged$x[,2],
  Condition = combined_labels
)

ggplot(pca_df_unmerged, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of Unmerged Data", x = "PC1", y = "PC2")
```
```{r}
# PCA on combat merged data (batch corrected)
pca_combat <- prcomp(t(combat_corrected_expr), scale. = TRUE)
pca_df_combat <- data.frame(
  PC1 = pca_combat$x[,1],
  PC2 = pca_combat$x[,2],
  Condition = combined_labels,
  Batch = batch_labels
)

# Plot 2: PCA of corrected data, colored by condition
ggplot(pca_df_combat, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of ComBat-Corrected Data (by Condition)", x = "PC1", y = "PC2")

```

```{r}
# PCA of merged data, coloured by batch
ggplot(pca_df_combat, aes(x = PC1, y = PC2, color = Batch)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of ComBat-Corrected Data (by Dataset)", x = "PC1", y = "PC2")

```

4 types of models (performance based on accuracy)
```{r}
set.seed(123)

# Store results
all_results <- data.frame()

# List of top N genes
for (n in c(2, 5, 10, 20)) {
  top_genes <- rownames(top_genes_df)[1:n]
  expr_subset <- t(expr_non_healthy[top_genes, ])
  
  df <- data.frame(expr_subset, label = factor(labels_non_healthy, levels = c("moderate", "severe")))
  
  # Set up CV
  ctrl <- trainControl(method = "cv", number = 5, savePredictions = TRUE)
  
  # Models to train
  models <- c("glm", "svmRadial", "rf", "knn")
  
  for (model in models) {
    fit <- train(label ~ ., data = df, method = model, trControl = ctrl)
    
    # Extract accuracy for each fold
    accs <- fit$resample$Accuracy
    
    # Store results in long format
    all_results <- rbind(all_results, data.frame(
      Accuracy = accs,
      Model = model,
      TopGenes = paste0("Top_", n)
    ))
  }
}

```

```{r}
all_results <- all_results %>%
  mutate(
    GeneModel = paste(TopGenes, Model, sep = "_"),
    Model = factor(Model, levels = c("glm", "knn", "rf", "svmRadial")),
    TopGenes = factor(TopGenes, levels = c("Top_2", "Top_5", "Top_10", "Top_20"))
  )

simplified_labels <- rep(c("GLM", "KNN", "RF", "SVM"), times = length(levels(all_results$TopGenes)))


# Order the new x-axis factor so the gene sets are grouped, and within them, model order is preserved
gene_model_levels <- c()
for (g in levels(all_results$TopGenes)) {
  for (m in levels(all_results$Model)) {
    gene_model_levels <- c(gene_model_levels, paste(g, m, sep = "_"))
  }
}
all_results$GeneModel <- factor(all_results$GeneModel, levels = gene_model_levels)

# Step 2: Plot with x = GeneModel
ggplot(all_results, aes(x = GeneModel, y = Accuracy, color = TopGenes)) +
  geom_boxplot(fill = NA, outlier.shape = NA) +
  geom_point(size = 2, alpha = 0.8, position = position_jitter(width = 0.15, height = 0)) +
  scale_x_discrete(labels = simplified_labels) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Secondary Model Accuracy for Top 2-20 DE Genes",
    x = "Model",
    y = "Accuracy",
    color = "Gene Set"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
4 model types (performance based on precision)

```{r}
install.packages("yardstick")
```


```{r}
library(caret)
library(ggplot2)
library(dplyr)
library(yardstick)

set.seed(123)

# Custom summary function to compute precision
precision_summary <- function(data, lev = NULL, model = NULL) {
  precision_val <- precision_vec(truth = data$obs, estimate = data$pred, event_level = "second")
  out <- c(Precision = precision_val)
  return(out)
}

# Store results
all_results <- data.frame()

# List of top N genes
for (n in c(2, 5, 10, 20)) {
  top_genes <- rownames(top_genes_df)[1:n]
  expr_subset <- t(expr_non_healthy[top_genes, ])
  
  df <- data.frame(expr_subset, label = factor(labels_non_healthy, levels = c("moderate", "severe")))
  
  # CV with precision as metric
  ctrl <- trainControl(
    method = "cv",
    number = 5,
    savePredictions = TRUE,
    classProbs = FALSE,
    summaryFunction = precision_summary
  )
  
  models <- c("glm", "svmRadial", "rf", "knn")
  
  for (model in models) {
    fit <- train(label ~ ., data = df, method = model, metric = "Precision", trControl = ctrl)
    
    # Extract precision values
    precisions <- fit$resample$Precision
    
    # Store results
    all_results <- rbind(all_results, data.frame(
      Precision = precisions,
      Model = model,
      TopGenes = paste0("Top_", n)
    ))
  }
}

# Organize labels and factors for plotting
all_results <- all_results %>%
  mutate(
    GeneModel = paste(TopGenes, Model, sep = "_"),
    Model = factor(Model, levels = c("glm", "knn", "rf", "svmRadial")),
    TopGenes = factor(TopGenes, levels = c("Top_2", "Top_5", "Top_10", "Top_20"))
  )

# Simplified model labels (repeated pattern)
simplified_labels <- rep(c("GLM", "KNN", "RF", "SVM"), times = length(levels(all_results$TopGenes)))

# Preserve grouping and order for x-axis
gene_model_levels <- c()
for (g in levels(all_results$TopGenes)) {
  for (m in levels(all_results$Model)) {
    gene_model_levels <- c(gene_model_levels, paste(g, m, sep = "_"))
  }
}
all_results$GeneModel <- factor(all_results$GeneModel, levels = gene_model_levels)

# Precision plot
ggplot(all_results, aes(x = GeneModel, y = Precision, color = TopGenes)) +
  geom_boxplot(fill = NA, outlier.shape = NA) +
  geom_point(size = 2, alpha = 0.8, position = position_jitter(width = 0.15, height = 0)) +
  scale_x_discrete(labels = simplified_labels) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Secondary Model Precision with Top 2-100 DE Genes",
    x = "Model",
    y = "Precision",
    color = "Gene Set"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



## 1-step model comparison

```{r}
set.seed(123)

# Custom summary function to return both Accuracy and macro-averaged Precision
multi_class_metrics <- function(data, lev = NULL, model = NULL) {
  acc <- yardstick::accuracy_vec(data$obs, data$pred)
  
  # Average precision across all classes (macro average)
  precision_per_class <- data %>%
    yardstick::precision(truth = obs, estimate = pred) %>%
    dplyr::pull(.estimate)
  
  macro_precision <- mean(precision_per_class)
  
  c(Accuracy = acc, Precision = macro_precision)
}

# Storage
all_results_1step <- data.frame()

# Top 5 genes
top_genes_5 <- rownames(top_genes_df)[1:5]
expr_subset_1step <- t(combat_corrected_expr[top_genes_5, ])
df_1step <- data.frame(expr_subset_1step, label = factor(combined_labels, levels = c("healthy", "moderate", "severe")))

# Cross-validation
ctrl_1step <- trainControl(
  method = "cv",
  number = 5,
  savePredictions = TRUE,
  classProbs = FALSE,
  summaryFunction = multi_class_metrics
)

models <- c("svmRadial", "rf", "knn")

# Train models and collect metrics
for (model in models) {
  fit <- train(label ~ ., data = df_1step, method = model, metric = "Accuracy", trControl = ctrl_1step)
  
  accs <- fit$resample$Accuracy
  precs <- fit$resample$Precision
  
  all_results_1step <- rbind(all_results_1step, data.frame(
    Accuracy = accs,
    Precision = precs,
    Model = model
  ))
}

```

```{r}
# Boxplot to compare the accuracy of different models
ggplot(all_results_1step, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Model Accuracy Comparison (5-Fold CV)",
    x = "Model Type",
    y = "Accuracy (5-fold CV)"
  ) +
  scale_fill_manual(values = c("svmRadial" = "lightblue", "rf" = "lightgreen", "knn" = "lightcoral")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
ggplot(all_results_1step, aes(x = Model, y = Precision, fill = Model)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Model Precision Comparison (Macro Avg, 5-Fold CV)",
    x = "Model Type",
    y = "Macro-Averaged Precision"
  ) +
  scale_fill_manual(values = c("svmRadial" = "lightblue", "rf" = "lightgreen", "knn" = "lightcoral")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
all_results_1step %>%
  group_by(Model) %>%
  summarise(
    MeanAccuracy = mean(Accuracy),
    MeanPrecision = mean(Precision)
  ) %>%
  arrange(desc(MeanPrecision))

```


## 2-step model comparative accuracy

```{r}
# Set the labels for binary classification: "asthma" vs "healthy"
binary_labels <- ifelse(combined_labels != "healthy", "asthma", "healthy")

# Subset the expression data for the top 10 genes
top_genes_10 <- rownames(top_genes_df)[1:10]
expr_subset_10 <- t(combat_corrected_expr[top_genes_10, ])

# Create a data frame with the expression data and binary labels
df_binary_2step <- data.frame(expr_subset_10, label = factor(binary_labels, levels = c("healthy", "asthma")))
```

```{r}
# Set up cross-validation (5-fold CV)
ctrl_binary <- trainControl(method = "cv", number = 5, savePredictions = TRUE)

# Train the logistic regression model
logit_model <- train(label ~ ., data = df_binary_2step, method = "svmRadial", trControl = ctrl_binary)

# Print the model results
cat("Logistic Regression Model Accuracy (Asthma vs Healthy): ", logit_model$results$Accuracy, "\n")
```

```{r}
# --- STEP 1: Get top genes and prepare data for step 1 (healthy vs asthma) ---
top_genes_step1 <- rownames(top_genes_df)[1:20]
expr_step1 <- t(combat_corrected_expr[top_genes_step1, ])
binary_labels <- ifelse(combined_labels == "healthy", "healthy", "asthma")
df_step1 <- data.frame(expr_step1, label = factor(binary_labels, levels = c("healthy", "asthma")))

# Train final Step 1 model: RF
set.seed(123)
step1_model <- train(
  label ~ ., data = df_step1,
  method = "rf",
  trControl = trainControl(method = "none")  # train on full data
)

# --- STEP 2: Get top genes and prepare data for step 2 (moderate vs severe) ---
top_genes_step2 <- rownames(top_genes_df)[1:5]  # top 5 DE genes for moderate vs severe
expr_step2 <- t(expr_non_healthy[top_genes_step2, ])
df_step2 <- data.frame(expr_step2, label = factor(labels_non_healthy, levels = c("moderate", "severe")))

# Train final Step 2 model: kNN
set.seed(123)
step2_model <- train(
  label ~ ., data = df_step2,
  method = "knn",
  trControl = trainControl(method = "none")
)

# --- STEP 3: Make predictions ---
# Predict Step 1 (asthma vs healthy)
pred_step1 <- predict(step1_model, newdata = df_step1)

# Prepare Step 2 data: only predict moderate/severe if Step 1 says "asthma"
step2_input <- combat_corrected_expr[top_genes_step2, pred_step1 == "asthma", drop = FALSE]
step2_input <- t(step2_input)
step2_input_df <- data.frame(step2_input)

# Predict Step 2 (moderate vs severe)
pred_step2 <- predict(step2_model, newdata = step2_input_df)

# --- STEP 4: Combine Step 1 and Step 2 predictions ---
# Construct final prediction: if step 1 == healthy, predict "healthy", otherwise use step 2
final_preds <- rep(NA, length(pred_step1))
final_preds[pred_step1 == "healthy"] <- "healthy"
final_preds[pred_step1 == "asthma"] <- as.character(pred_step2)

# --- STEP 5: Compare to true labels ---
# True labels are in combined_labels
final_preds <- factor(final_preds, levels = c("healthy", "moderate", "severe"))
true_labels <- factor(combined_labels, levels = c("healthy", "moderate", "severe"))

# Compute overall accuracy
overall_accuracy <- mean(final_preds == true_labels)
cat("Overall 2-Step Model Accuracy:", overall_accuracy, "\n")

# Optional: confusion matrix
confusionMatrix(final_preds, true_labels)

```

```{r}
library(caret)
library(dplyr)

set.seed(123)
num_splits <- 5
accuracies <- numeric(num_splits)

# Prepare true class labels
true_labels_all <- factor(combined_labels, levels = c("healthy", "moderate", "severe"))

# Create binary asthma/healthy labels for step 1
binary_labels_all <- ifelse(combined_labels == "healthy", "healthy", "asthma")

# Loop through 5 random splits
for (i in 1:num_splits) {
  cat("Running split", i, "...\n")
  
  # Create train/test indices
  train_idx <- createDataPartition(true_labels_all, p = 0.8, list = FALSE)
  test_idx <- setdiff(seq_along(true_labels_all), train_idx)
  
  # --- STEP 1: Train RF model for asthma vs healthy ---
  expr_step1_train <- t(combat_corrected_expr[top_genes_step1, train_idx])
  df_step1_train <- data.frame(expr_step1_train,
                               label = factor(binary_labels_all[train_idx], levels = c("healthy", "asthma")))
  
  step1_model <- train(
    label ~ ., data = df_step1_train,
    method = "rf",
    trControl = trainControl(method = "none")
  )
  
  # --- STEP 2: Train kNN model for moderate vs severe ---
  # Only train on non-healthy samples in the training set
  non_healthy_train_idx <- train_idx[which(combined_labels[train_idx] != "healthy")]
  expr_step2_train <- t(combat_corrected_expr[top_genes_step2, non_healthy_train_idx])
  df_step2_train <- data.frame(expr_step2_train,
                               label = factor(combined_labels[non_healthy_train_idx],
                                              levels = c("moderate", "severe")))
  
  step2_model <- train(
    label ~ ., data = df_step2_train,
    method = "knn",
    trControl = trainControl(method = "none")
  )
  
  # --- STEP 3: Apply to test data ---
  # Step 1 prediction
  expr_step1_test <- t(combat_corrected_expr[top_genes_step1, test_idx])
  df_step1_test <- data.frame(expr_step1_test)
  pred_step1 <- predict(step1_model, newdata = df_step1_test)
  
  # Step 2: Only apply to test samples predicted as asthma
  asthma_test_idx <- test_idx[which(pred_step1 == "asthma")]
  
  if (length(asthma_test_idx) > 0) {
    expr_step2_test <- t(combat_corrected_expr[top_genes_step2, asthma_test_idx])
    df_step2_test <- data.frame(expr_step2_test)
    pred_step2 <- predict(step2_model, newdata = df_step2_test)
  } else {
    pred_step2 <- character(0)
  }
  
  # --- STEP 4: Final combined prediction ---
  final_preds <- character(length(test_idx))
  final_preds[pred_step1 == "healthy"] <- "healthy"
  final_preds[pred_step1 == "asthma"] <- as.character(pred_step2)
  
  # Final true labels
  true_test_labels <- combined_labels[test_idx]
  
  # --- STEP 5: Accuracy ---
  final_preds <- factor(final_preds, levels = c("healthy", "moderate", "severe"))
  true_test_labels <- factor(true_test_labels, levels = c("healthy", "moderate", "severe"))
  
  acc <- mean(final_preds == true_test_labels)
  accuracies[i] <- acc
}

# Final output
cat("Average Overall 2-Step Model Accuracy across", num_splits, "splits:", round(mean(accuracies), 4), "\n")

```
```{r}
set.seed(2)
num_splits <- 5
accuracies <- numeric(num_splits)
precisions <- numeric(num_splits)

# Prepare full labels
true_labels_all <- factor(combined_labels, levels = c("healthy", "moderate", "severe"))
binary_labels_all <- ifelse(combined_labels == "healthy", "healthy", "asthma")

for (i in 1:num_splits) {
  cat("Running split", i, "...\n")
  
  # Split
  train_idx <- createDataPartition(true_labels_all, p = 0.8, list = FALSE)
  test_idx <- setdiff(seq_along(true_labels_all), train_idx)
  
  # --- STEP 1: Train RF model (healthy vs asthma) ---
  expr_step1_train <- t(combat_corrected_expr[top_genes_step1, train_idx])
  df_step1_train <- data.frame(expr_step1_train,
                               label = factor(binary_labels_all[train_idx], levels = c("healthy", "asthma")))
  
  step1_model <- train(label ~ ., data = df_step1_train, method = "rf",
                       trControl = trainControl(method = "none"))
  
  # --- STEP 2: Train kNN model (moderate vs severe) ---
  non_healthy_train_idx <- train_idx[which(combined_labels[train_idx] != "healthy")]
  expr_step2_train <- t(combat_corrected_expr[top_genes_step2, non_healthy_train_idx])
  df_step2_train <- data.frame(expr_step2_train,
                               label = factor(combined_labels[non_healthy_train_idx],
                                              levels = c("moderate", "severe")))
  
  step2_model <- train(label ~ ., data = df_step2_train, method = "knn",
                       trControl = trainControl(method = "none"))
  
  # --- STEP 3: Predict on test set ---
  expr_step1_test <- t(combat_corrected_expr[top_genes_step1, test_idx])
  df_step1_test <- data.frame(expr_step1_test)
  pred_step1 <- predict(step1_model, newdata = df_step1_test)
  
  asthma_test_idx <- test_idx[which(pred_step1 == "asthma")]
  if (length(asthma_test_idx) > 0) {
    expr_step2_test <- t(combat_corrected_expr[top_genes_step2, asthma_test_idx])
    df_step2_test <- data.frame(expr_step2_test)
    pred_step2 <- predict(step2_model, newdata = df_step2_test)
  } else {
    pred_step2 <- character(0)
  }
  
  # --- STEP 4: Final prediction ---
  final_preds <- character(length(test_idx))
  final_preds[pred_step1 == "healthy"] <- "healthy"
  final_preds[pred_step1 == "asthma"] <- as.character(pred_step2)
  
  true_test_labels <- combined_labels[test_idx]
  
  # Format factors
  final_preds <- factor(final_preds, levels = c("healthy", "moderate", "severe"))
  true_test_labels <- factor(true_test_labels, levels = c("healthy", "moderate", "severe"))
  
  # --- STEP 5: Accuracy and Precision ---
  acc <- mean(final_preds == true_test_labels)
  accuracies[i] <- acc
  
  # Use yardstick for macro-averaged precision
  prec_tbl <- data.frame(truth = true_test_labels, prediction = final_preds)
  prec_result <- yardstick::precision(prec_tbl, truth = truth, estimate = prediction, estimator = "macro")
  precisions[i] <- prec_result$.estimate
}

# --- FINAL RESULTS ---
cat("Average 2-Step Model Accuracy over", num_splits, "splits:", round(mean(accuracies), 4), "\n")
cat("Average Macro-Averaged Precision over", num_splits, "splits:", round(mean(precisions), 4), "\n")

```



## Final kNN Secondary Model and RDS for Shiny

```{r}
set.seed(123)

# Select top 5 genes
top5_genes <- rownames(top_genes_df)[1:5]
expr_top5 <- t(expr_non_healthy[top5_genes, ])

df_final <- data.frame(expr_top5, label = factor(labels_non_healthy, levels = c("moderate", "severe")))

final_knn_model <- train(
  label ~ .,
  data = df_final,
  method = "knn",
  trControl = trainControl(method = "none")
)
```

RDS stuff

```{r}
saveRDS(combat_corrected_expr, file = "frozen_combat_corrected_expr_secondary_model.rds")

# --- 2. Save list of top DE genes for both steps ---
top_genes_secondary <- rownames(top_genes_df)[1:5] 
top_genes_secondary

saveRDS(top_genes_secondary, file = "top_de_genes_list_secondary.rds")

# --- 3. Save final kNN model for moderate vs severe ---
saveRDS(final_knn_model, file = "final_secondary_model.rds")

# Add gene names as a column



```

```{r}
combat_expr_export <- data.frame(Gene = rownames(expr_non_healthy), expr_non_healthy)
write.csv(combat_expr_export, file = "combat_combined_expression_matrix_secondary_model.csv", row.names = FALSE)

```

```{r}
expr_non_healthy_raw <- raw_expr_matrix[, non_healthy_idx]

raw_expr_export <- data.frame(Gene = rownames(expr_non_healthy_raw), expr_non_healthy_raw)
write.csv(raw_expr_export, file = "raw_combined_expression_matrix_secondary_model.csv", row.names = FALSE)
```






